---
title: "Introduction to OceanSDM"
date: 2026-07-21
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: false
    theme: cosmo
    highlight: tango
    css: vignette.css
vignette: >
  %\VignetteIndexEntry{Introduction to OceanSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Predicting the dynamic, three-dimensional distributions of marine species is challenging given most SDM tools currently can only predict static 2D habitats (but see an example of voluModel package,Owens & Rahbek, 2023). Here, we present a novel package 'OceanSDM' to fill this gap. This package includes 18 user-friendly functions helping researchers to complete tasks from initial data extraction to final publication-quality figure creation. It can be split into three major sections: (1) Prepare species and environmental data, (2) Build 2D SDMs for each temporal bin and generate related figures, and (3) Estimate suitable/preferred depth and map the 3D habitat for each temporal bin. Here I illustrate the utility of this package with a pelagic shark (whale shark). For the temporal bin, I used quarterly bins (four bins).

Here are the packages you need.
```{r setup}
# Install the package if you haven't done this yet.
# if (!require("remotes")) install.packages("remotes")
# remotes::install_github("HarryZhang0130/OceanSDM",dependencies = TRUE) # When you are asked to update dependent packages, you can skip all but make sure you installed the 'sdm' package (>= 1.2=59).
library(OceanSDM) # this package
library(sdm) # for using horizontal species distribution modelling functions
library(lubridate) # for date format
```
# 1. Prepare species and environmental data 
## 1.1 Download species data from GBIF and OBIS using ob_data() 
The Global Biodiversity Information Faculty (GBIF) and Ocean Biodiversity Information System (OBIS) are the two major sources of species occurrences at the global scale. The function ob_data() helps users automatically download the geo-referenced occurrences of either a single or many species. Here I demonstrated with an example of smaller temporal and spatial ranges than the real case in the paper to save the running time for creating this vignette.

  Here below are the parameters used in the ob_data().
  
@param key Character. The species scientific name (e.g., "Rhincodon typus").

@param sp_geometry Character. The geographic range of the study area as a WKT polygon
  (e.g., "POLYGON ((20 -41, 20 40, 180 40, 180 -41, 20 -41))").

@param start_date Character. Start date used in `robis::occurrence()`, format "YYYY-MM-DD".

@param end_date Character. End date used in `robis::occurrence()`, format "YYYY-MM-DD".

@param event_date Character. Year range used in `rgbif::occ_search()`, format "YYYY-MM-DD,YYYY-MM-DD".

```{r occurrence, dev='png', cache=TRUE}
 ws_occ<-ob_data(
    key<-'Rhincodon typus' ,# It can be a vector of the Latin names of all focal species
    sp_geometry <- "POLYGON ((90 -41, 90 40, 180 40, 180 -41, 90 -41))", # Replace this to "POLYGON ((90 -41, 90 40, 180 40, 180 -41, 90 -41))" to simulate my real case study!!!
    start_date<-"2015-01-01", # Replace this to "2015-01-01" to simulate my real case study!!!
    end_date<-"2024-12-30",
    event_date<-'2015-01-01,2024-12-30') # Replace this to '2015-01-01,2024-12-30' to simulate my real case study!!!
# # save the occurrences
write.csv(ws_occ$sp_occ,"F:/whaleshark_sdm/occ/R_typus.csv",row.names = F)
```

## 1.2 Download .nc files of marine environmental predictors using env_download() 
The Copernicus Marine Service (CMS) offers dynamic, three-dimensional environmental data that can be extracted with the function env_download(). !!! But please note you must first get access to the copernicusmarine toolbox on which this function was built.
Here is a step-by-step instruction for this:

Step 1. Download the executable file

If you are the first time to use copernicusmarine toolbox. You should first download an executable file of the copernicusmarine toolbox. See the download links in the “Step 2” of an official instruction at https://help.marine.copernicus.eu/en/articles/10750437-copernicus-marine-toolbox-executable-no-installation. For different OS (Windows, Mac, Linux), there are different executable files. You should choose the right one.

Step 2. Move it to a safe directory and register

After download it, make sure you move it to a directory (e.g., in my Windows, I stored it at “C:”) that you won’t clean when you utilize storage cleaning up software. If you haven’t registered to the Copernicus Marine Service yet, you can register at https://data.marine.copernicus.eu/register.

Step 3. Login to the Copernicus Marine Service

For Windows users, go to your command prompt and navigate to the executable file. For instance, if you are using Windows and the executable file is stored at “C:/shark”, then in the command prompt, run this line to move to your directory: 
cd “C:/shark”. Then run the login code: copernicusmarine login

This will allow you to use your registered username and password to create a credential file (.copernicusmarine-credentials) and store it in your PC. 

For those who use MacOS or Linux, you need an extra step ahead to grant execute permission. Suppose you downloaded the executable file (copernicusmarine_macos-x86_64.cli) and stored the executable file in a safe path (e.g., ./shark). Open the terminal in this path, then run: chmod +x copernicusmarine_macos-x86_64

After this you can log in by runnig this line:./copernicusmarine_macos-x86_64 login

Now, you are ready to download the required product in R by using the env_download() function. 


  Here below are the parameters used in the env_download().
  
@param path_cmems_tool Character. Path to the copernicusmarine executable,
  e.g. `"/miniconda/envs/copernicusmarine/Scripts/copernicusmarine.exe"`.

@param out_dir Character. Directory where output data will be stored.

@param productId Character. ID of the CMEMS product (see dataset user manual
  in the CMEMS product webpage). e.g., `"cmems_mod_glo_phy-mnstd_my_0.25deg_P1M-m"`.

@param variable Character. Variable name to download. e.g., "thetao_mean".

@param date_min Date or character. Minimum date. e.g., "1993-01-01".

@param date_max Date or character. Maximum date. e.g., "2022-12-01".

@param lon Numeric vector of length 2. Longitude range (min, max).
  e.g., `c(-180, 179.75)`.

@param lat Numeric vector of length 2. Latitude range (min, max).
  e.g., `c(-80, 90)`.

@param depth Numeric vector of length 2. Depth range in meters (min, max).
  e.g., `c(0.51, 199.79)`.

@param out_name Character. Output file name ending with `.nc`.
```{r set up tool path, cache=TRUE}
## Set up the path of the cms downloading tool and output directory
path_cms<-"C:/shark/miniconda/envs/copernicusmarine/Scripts/copernicusmarine.exe" # replace to your own directory
out_dir <- "F:/whaleshark_sdm/cmems" # replace to your own directory

```
### 1.2.1 Extract surface layer of marine environmental predictors
```{r download surface predictor data, cache=TRUE}
## Note here is a simplified example for fast creating the vignettes
productId <- "cmems_mod_glo_phy-mnstd_my_0.25deg_P1M-m"
variable <- " --variable thetao_mean"
date_min <- ymd(20240101) # start_date, it should be ymd(20150101) in the paper
date_max <- ymd(20241201) # end_date
lon<-c(90,180) # longitudinal range
lat<-c(-41,40) # latitudinal range
## sea surface temperature
depth <- list(0.51, 2) # depth_min, depth_max
out_name <- paste("IndoWPac_temp_1m_y2024.nc", sep="") # it should be paste("IndoWPac_temp_1m_y2015_y2024.nc", sep="") in the paper
env_download(path_cms,out_dir,productId,variable,
              date_min,date_max,lon,lat,depth,out_name)
## sea surface height
# variable <- " --variable zos_mean"
# out_name <- paste("IndoWPac_slh_1m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
## mixed layer depth
# variable <- " --variable mlotst_mean"
# out_name <- paste("IndoWPac_mlt_1m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
## salinity
# variable <- " --variable so_mean"
# out_name <- paste("IndoWPac_so_1m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
```

### 1.2.2 Extract multiple depth layers of ocean temperature
This step is NOT necessary for the 2D SDM construction, but is required for estimating suitable/tolerable depths based on species realized climatic niche and the three-dimensional distributions of the focal climatic variable (here, temperature) over the chosen temporal bins (here, four quarters). Here given the size of the multi-depth layers of data, I split it into four .nc files (0-200m, 200-400m, 400-1050m, 1050-1946m). The maximum depth (1946m) was based on the known depth records of the whale shark in the study area. The codes were set to 'not run' with the hash tag while creating this vignette as it was time consuming. Users can simply remove these hash tags (Using 'Control+Shift+C' for Windows or 'Command+Shift+C" for Mac) and run the code.
```{r download multi-depth layers of temperature, cache=TRUE}
# variable <- " --variable thetao_mean"
# out_dir = "F:/whaleshark_sdm/cmems"
# productId = "cmems_mod_glo_phy-mnstd_my_0.25deg_P1M-m"
# date_min = ymd(20150101) # start_date
# date_max = ymd(20241201) # end_date
# lon<-c(90,180)
# lat<-c(-41,40)
# depth <- list(0.50, 200) # depth_min, depth_max
# out_name <- paste("IndoWPac_temp_200m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
# depth <- list(200, 400) # depth_min, depth_max
# out_name <- paste("IndoWPac_temp_200_400m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
# depth <- list(400, 1050) # depth_min, depth_max
# out_name <- paste("IndoWPac_temp_400_1050m_y2015_y2024.nc", sep="")
#env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
# depth <- list(1050, 1946) # depth_min, depth_max
# out_name <- paste("IndoWPac_temp_1050_1946m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
```

### 1.2.3 Extract sea surface data for net primary productivity and O2 from another CMS dataset
```{r download npp and o2, cache=TRUE}
# out_dir <- "F:/cmems"
# productId <- "cmems_mod_glo_bgc_my_0.25deg_P1M-m"
# date_min <- ymd(20150101) # start_date
# date_max <- ymd(20241201) # end_date
# lon<-c(90,180)
# lat<-c(-41,40)
# depth <- list(0.51, 2) # depth_min, depth_max
# variable <- " --variable nppv"
# out_name <- paste("IndoWPac_npp_1m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)

# variable <- " --variable o2"
# out_name <- paste("IndoWPac_o2_1m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
```
For extracting predictor datasets of future projections, please check https://esgf-node.llnl.gov/search/cmip6/
There is also an R package for downloading CMIP6 data: https://github.com/marine-ecologist/CMIP6/blob/main/CMIP6_projections.R

### 1.2.4 Prepare temporally-binned predictors using env_bin()
To predict the dynamic distributions of marine species across the 2D space, all predictors should be aggregated into mean values within the specific temporal bins.  

  Here below are the parameters used in the env_bin().
  
@param sp_trait Character. Species vertical habitat use. One of `"pelagic"`,
  `"benthic"`, or `"benthopelagic"`.

@param env_path Character. Path to the downloaded CMEMS NetCDF file.

@param out_dir Character. Directory where the temporally‑binned raster layers
  will be saved.

@param var_name Character. Name of the variable (refer to CMEMS user manual).

@param lon_min Numeric. Minimum longitude (western boundary).

@param lon_max Numeric. Maximum longitude (eastern boundary).

@param lat_min Numeric. Minimum latitude (southern boundary).

@param lat_max Numeric. Maximum latitude (northern boundary).

@param date_min Date (or character convertible to Date). Start date.

@param date_max Date (or character convertible to Date). End date.

@param lag_days Numeric. Number of days ocean seasonality lags behind
  atmospheric seasonality. Default `76` if missing.

@param temporal_bin Character. How a year will be split. One of `"monthly"`,
  `"quarterly"`, or `"seasonal"`.
```{r env_bin, cache=FALSE}
gc() # clear the R environment 
## sea surface temperature (Note this is a simplified example of using downloaded data in 1.2.1)
env_bin(sp_trait="pelagic",            
        env_path="F:/whaleshark_sdm/cmems/IndoWPac_temp_1m_y2024.nc", 
# in the paper, I used "F:/whaleshark_sdm/cmems/IndoWPac_temp_1m_y2015_y2024.nc"
       out_dir="F:/whaleshark_sdm/bin2",  
# in the paper, I used "F:/whaleshark_sdm/bin". here I set the directory to a temporal folder to receive the output from this example.
        var_name="thetao_mean",
        lon_min=90,
       lon_max=180,
        lat_min=-41,
        lat_max=40, # in the paper, I used 40
       date_min=ymd(20240101), # in the paper, I used ymd(20150101)
        date_max = ymd(20241201),
        temporal_bin="quarterly")
## sea surface height
# env_bin(sp_trait="pelagic",
#        env_path="F:/cmems/IndoWPac_slh_1m_y2015_y2024.nc",
#        out_dir="F:/whaleshark_sdm/bin",
#        var_name="zos_mean",
#        lon_min=90,
#        lon_max=180,
#        lat_min=-41,
#        lat_max=40,
#        date_min=ymd(20150101),
#        date_max = ymd(20241201),

#        temporal_bin="quarterly")
## mixed layer depth
# env_bin(sp_trait="pelagic",
#        env_path="F:/cmems/IndoWPac_mlt_1m_y2015_y2024.nc",
#        out_dir="F:/whaleshark_sdm/bin",
#        var_name="mlotst_mean",
#        lon_min=90,
#        lon_max=180,
#        lat_min=-41,
#        lat_max=40,
#        date_min=ymd(20150101),
#        date_max = ymd(20241201),

#        temporal_bin="quarterly")
## sea surface salinity
# env_bin(sp_trait="pelagic",
#        env_path="F:/cmems/IndoWPac_so_1m_y2015_y2024.nc",
#        out_dir="F:/whaleshark_sdm/bin",
#        var_name="so_mean",
#        lon_min=90,
#        lon_max=180,
#        lat_min=-41,
#        lat_max=40,
#        date_min=ymd(20150101),
#        date_max = ymd(20241201),

#        temporal_bin="quarterly")
## sea surface NPP
# env_bin(sp_trait="pelagic",
#        env_path="F:/cmems/IndoWPac_npp_1m_y2015_y2024.nc",
#        out_dir="F:/whaleshark_sdm/bin",
#        var_name="nppv",
#        lon_min=90,
#        lon_max=180,
#        lat_min=-41,
#        lat_max=40,
#        date_min=ymd(20150101),
#        date_max = ymd(20241201),

#        temporal_bin="quarterly")
## sea surface dissolved oxygen
# env_bin(sp_trait="pelagic",
#        env_path="F:/cmems/IndoWPac_o2_1m_y2015_y2024.nc",
#        out_dir="F:/whaleshark_sdm/bin",
#        var_name="o2",
#        lon_min=90,
#        lon_max=180,
#        lat_min=-41,
#        lat_max=40,
#        date_min=ymd(20150101),
#        date_max = ymd(20241201),

#        temporal_bin="quarterly")
```
Please note that env_bin() function will create sub-folders (e.g., Q1, Q2, Q3, Q4) for each temporal bin, and save the predictors to the corresponding sub-folders. You may, however, need to rename them to the more meaningful type if necessary. For instance, ocean temperature is usually named as "thetao_mean" in the original cms dataset, and you may change it to "Temperature". 

### 1.2.5 Download the depth dataset 
The Bio-ORACLE provides statistics of ocean depths (mean, maximum, etc.). Here I did not run the code while creating this vignette, but you can run it after removing hash tags. 
```{r download depth, cache=TRUE}
# devtools::install_github("bio-oracle/biooracler")
# library(biooracler)
# library(terra)
# list<-list_layers("Ocean depth")
# dataset_id <- list$dataset_id[158]
# latitude <- c(-41, 40)
# longitude <- c(90, 179.975)
# constraints <- list(latitude, longitude)
# names(constraints) <- c("latitude", "longitude")
# variables <- "bathymetry_mean"
# depth <- download_layers(dataset_id, variables, constraints)
# depth<-abs(depth) # convert values to positive.
# temp<-terra::rast("F:/whaleshark_sdm/bin/Q1/Temperature.asc")
# depth_resampled <- terra::resample(depth, temp, method = "bilinear")
# writeRaster(depth_resampled,"F:/whaleshark_sdm/bin/Depth.asc",
#             NAflag=-9999,overwrite=TRUE,filetype = "AAIGrid")
```
Please note that you need to add the depth_mean.asc sub-folders (e.g., Q1, Q2, Q3, Q4) for each temporal bin, and save the predictors to the corresponding sub-folders. You may, however, need to rename them to the more meaningful type if necessary. For instance, ocean temperature is usually named as "thetao_mean" in the original cms dataset, and you may change it to "temperature". I have uploaded the prepared quarterly predictor datasets to the package, so I will get access to these system files directly as follows. 

### 1.2.6 Plot environmental predictors (optional)
```{r plot_data, cache=TRUE}
# Load required packages
# library(terra)
# library(ggplot2)
# library(cowplot)
# library(dplyr)
# library(scales)   # for squish
# 
# # Main directory and quarters
# main_dir <- system.file("extdata", package = "OceanSDM") # replace to your own directory where the predictor data were stored
# quarters <- c("Q1", "Q2", "Q3", "Q4") # replace to the sub-folder for each quarter under the main_dir
# 
# # Loop over each quarter
# for (q in quarters) {
#   quarter_path <- file.path(main_dir, q)
# 
#   # Read all .asc files in the quarter folder
#   files <- list.files(quarter_path, pattern = "\\.asc$", full.names = TRUE,
#              ignore.case = TRUE)
#   r <- terra::rast(files)
#   names(r) <- tools::file_path_sans_ext(basename(files))
# 
#   n_layers <- nlyr(r)
# 
# ## Convert raster layers to a data frame for ggplot
#   df_list <- list()
#   for (i in 1:n_layers) {
#     df <- as.data.frame(r[[i]], xy = TRUE, na.rm = FALSE)
#     colnames(df) <- c("x", "y", "value")
#     df$variable <- names(r)[i]
#     df_list[[i]] <- df
#   }
#   df_all <- bind_rows(df_list)
# 
# # Create a list of ggplot objects with external labels
#   gg_list <- list()
#   for (i in 1:n_layers) {
#     layer_name <- names(r)[i]
#     df_sub <- filter(df_all, variable == layer_name)
# ## Compute 2% and 98% percentiles for color stretching (percent clip)
#     values_vec <- df_sub$value[!is.na(df_sub$value)]
#     if (length(values_vec) > 0) {
#       p2 <- quantile(values_vec, 0.02, na.rm = TRUE)
#       p98 <- quantile(values_vec, 0.98, na.rm = TRUE)
#       # If p2 == p98, avoid identical limits; use full range
#       if (p2 == p98) {
#         p2 <- min(values_vec, na.rm = TRUE)
#         p98 <- max(values_vec, na.rm = TRUE)
#       }
#     } else {
#       p2 <- 0; p98 <- 1  # fallback
#     }
# 
#     p <- ggplot(df_sub, aes(x = x, y = y, fill = value)) +
#       geom_raster() +
#       scale_fill_viridis_c(
#         option = "inferno",
#         begin = 0.1,
#         end = 0.9,
#         limits = c(p2, p98),      # percent clip range
#         oob = squish,            # squish extremes to the nearest limit color
#         na.value = "transparent"
#       ) +
#       coord_fixed() +
#       labs(title = layer_name, fill = "Value") +
#       theme_minimal() +
#       theme(
#         plot.title = element_text(size =8, face = "bold", hjust = 0.5),
#         legend.position = "right",
#         legend.key.size = unit(0.25, "cm"),
#         legend.text = element_text(size = 4),
#         legend.title = element_text(size = 5),
#         legend.spacing = unit(0.1, "cm"),
#         axis.text = element_blank(),
#         axis.title = element_blank(),
#         axis.ticks = element_blank(),
#         panel.grid = element_blank(),
#         plot.margin = ggplot2::margin(t = 5, r = 5, b = 5, l = 5, unit = "pt")
#       )
# 
#     ggdraw_obj <- ggdraw(p) +
#       draw_label(
#         label = paste0("(", letters[i], ")"),
#         x = 0, y = 1,
#         hjust = -0.05,
#         vjust = 1,
#         size = 8,
#         fontface = "bold"
#       )
#     gg_list[[i]] <- ggdraw_obj
#   }
# 
# ## Arrange subplots
#   ncol <- ceiling(sqrt(n_layers))
#   nrow <- ceiling(n_layers / ncol)
#   combined <- plot_grid(plotlist = gg_list,
#                  ncol = ncol, nrow = nrow, align = "hv")
# 
# ## Create title with "Environmental predictors of QX" and separate by ~2 lines
#   title_text <- paste0("Environmental predictors of ", q)
#   title_gg <- ggdraw() +
#     draw_label(title_text, x = 0.5, y = 0.5, size = 16, fontface = "bold") +
#     theme(plot.margin = ggplot2::margin(b = 20, t = 10, unit = "pt"))
# 
# #   # Combine title and plot vertically
#   final_plot <- plot_grid(title_gg, combined, ncol = 1,
#                   rel_heights = c(0.12, 1), align = "v")
# 
# ## Display the plot
#   print(final_plot)
# }

```
## 1.3 Check occurrence coordinates against species range map with range_check_single() (optional) #############
Sometimes, you may need to check if all your species occurrences fall within the species range map. You can use range_check_single() for a single species, or range_check() for multiple taxa.

  Here below are the parameters used in the range_check_single().
 
@param species_name Character. Scientific name of the species.

@param occ_data Either a data frame with columns `decimalLongitude`, `decimalLatitude`,
  or a character string path to a CSV file containing those columns.

@param range_path Character. Path to the range map shapefile (must contain a column `sci_name`).

@param buffer_km Numeric. Buffer distance (km) around the range map to consider
  points as valid.

@param plot Logical. If `TRUE`, generate a map showing point categories. Default `FALSE`.

@param plot_file Character. Optional file path to save the plot as TIFF (e.g., "plot.tif").
  If `NULL`, the plot is displayed in the graphics device.

@param xlim Numeric vector of length 2 (min, max) for longitude limits. If `NULL`,
  automatically determined from the buffered range.

@param ylim Numeric vector of length 2 (min, max) for latitude limits. If `NULL`,
  automatically determined from the buffered range.
```{r range check, cache=FALSE}
# You can get access to the example occurrences of whale sharks in the system file folder extdata.
occ_path <-system.file("extdata", "R_typus.csv", package = "OceanSDM")

# You can get access to the example range map of whale sharks in the system file folder extdata.
range_path <-system.file("extdata", "R_typus_wip2.shp", package = "OceanSDM")
# But for creating this vignette, I loaded them from my PC.
#occ_path<-"F:/whaleshark_sdm/occ/R_typus.csv"
#range_path<-"F:/whaleshark_sdm/range_map/R_typus_wip2.shp"
result <- range_check_single(
  species_name="Rhincodon typus",
      occ_data = occ_path,
      range_path = range_path,
      buffer_km = 100,
  plot = T,
  xlim = c(90, 179.975),
  ylim = c(-41, 40)
  )
occ_range<-rbind(result$valid_points,result$invalid_points) # Note here I kept the 'invalid_points' because they are actually valid seasonal sightings even though they fell beyond the IUCN range map.
# # save the occurrences
write.csv(occ_range,"F:/whaleshark_sdm/occ/R_typus_r.csv",row.names = F)
```
## 1.4 Adjust occurrence coordinates with correct_ob() #############
Previous SDM studies often find that some occurrences were not spatially matched with the environmental layer, but these points fall within a close distance to the environmental layer. This may result from the resolution uncertainty of the occurrences or the coarse resolution of environmental data. I might be worthy correcting them so they match each other, especially when your focal species are naturally rare and you want to keep as many occurrences as possible. 

  Here below are the parameters used in the correct_ob().
 
@param occ_path Character. Path to occurrence CSV file. Must contain
  columns `decimalLongitude` and `decimalLatitude` (WGS84 coordinates).

@param predictor_raster_path Character. File path to the standard raster layer
  (e.g., an ASCII or GeoTIFF file) used as reference for correction.

@param max_distance_km Numeric. Maximum distance (in kilometres) that a point
  will be adjusted to the nearest grid cell with non-NA value.

@param out_dir Character. **Full file path** (including `.csv` extension) where
  the corrected occurrence data will be saved. If the directory does not exist,
  it will be created. If a directory path is given instead, a default filename
  `corrected_occurrences.csv` will be appended.

@param verbose Logical. If `TRUE`, print progress messages. Default is `TRUE`.

@param plot Logical. If `TRUE`, generate a map showing correction status. Default `FALSE`.

@param plot_file Character. Optional file path to save the plot as TIFF (e.g., "plot.tif"). If `NULL`, the plot is displayed in the graphics device.

@param xlim Numeric vector of length 2 (min, max) for longitude limits. If `NULL`,
  automatically determined from the raster extent.

@param ylim Numeric vector of length 2 (min, max) for latitude limits. If `NULL`,
  automatically determined from the raster extent.

```{r adjust_coordinates, cache=FALSE}

## Define input paths
ev_layer<-system.file("extdata", "Q1/Depth.asc", package = "OceanSDM")
occ_range<-system.file("extdata", "R_typus_r.csv", package = "OceanSDM")
output_path<-"F:/whaleshark_sdm/occ/R_typus_adj.csv"
corrected_ob <- correct_ob(
  occ_path = occ_range,
  predictor_raster_path = ev_layer,
  max_distance_km = 56,
  out_dir = output_path,
  verbose = TRUE, plot=T,
  xlim = c(90, 179.975),
  ylim = c(-41, 40)
)
summary(corrected_ob)

```

## 1.5 Check and remove spatio-temporal bias with bias_check() and bias_reduce()
Spatio-temporal sampling bias is a common issue in global biodiversity data and it degrades model robustness. Here, bias_check() and bias_reduce() are two functions to examine and reduce sampling bias (see details in Supplementary Information), respectively, adapted from the dynamicSDM (Dobson et al., 2023). 

  Here below are the parameters used in the bias_check().

@param occ_data Data frame. Species occurrence records. Must contain columns:
  `Species`, `x`, `y`, `year`, `month`, `day`.

@param temporal_level Character. Temporal level at which temporal sampling
  bias is to be tested. One of `month`, `quarter`, or `season`.

@param lag_days Numeric. Number of days ocean seasonality lags behind.

@param spatial_method Character. Method used to generate species habitat area
  for spatial random sampling. One of `convex_hull` or `range_map`.

@param world_path Character. File path to a world map shapefile (required when
  `spatial_method = "convex_hull"`).

@param range_path Character. File path to a species range map shapefile
  (required when `spatial_method = "range_map"`). Must contain a column
  `sci_name`.

@param prj Character. Coordinate reference system to be used for spatial maps,
  given as a PROJ string (e.g., "+proj=longlat +datum=WGS84").



  Here below are the parameters used in the bias_reduce().

@param occ_data Data frame. Species occurrence records. Must contain columns:
  "species", "x", "y", "year", "month", "day".

@param bias_result "BiasCheckResult" object. The result object returned by bias_check.

@param min_dist_km Numeric. Minimum distance (in kilometres) used for spatial thinning.

@param thin_reps Integer. Number of replicate thinning runs.

@param temporal_level Character. Temporal level used to address temporal
  sampling bias. Must be one of "month", "quarter", or "season".

@param seed Integer. Random seed for reproducibility of thinning and
  subsampling steps.

```{r check & remove bias, cache=TRUE}
library(dplyr)
## Read input data or skip this line if you have run the section 1.4
corrected_ob<-read.csv(system.file("extdata", "R_typus_adj.csv", package = "OceanSDM"))
## Read input data or skip this line if you have run the section 1.4
range_path <-system.file("extdata", "R_typus_wip2.shp", package = "OceanSDM")
colnames(corrected_ob)
occ_ob<-corrected_ob[,c(1:7)] # removed unuseful columns
colnames(occ_ob)[2:3]<-c("y","x") # rename latitude and longitude columns
#### For the quarterly SDMs
occ_ob2 <- occ_ob %>%
  mutate(
         quarter = dplyr::case_when(
           month %in% 1:3   ~ "Q1",
           month %in% 4:6   ~ "Q2",
           month %in% 7:9  ~ "Q3",
           month %in% 10:12 ~ "Q4",
           TRUE ~ NA_character_
         ))%>% dplyr::filter(!is.na(quarter))
table(occ_ob2$quarter)

#### For the seasonal SDMs
# occ_ob3 <- occ_ob2 %>%
#  mutate(date = make_date(year, month, day),
#         adjusted_date = date - days(73), 
# # Parse date information and adjust for sea temperature lag (73 days)
#         adjusted_month = month(adjusted_date),
#         hemisphere = ifelse(y >= 0, "Northern", "Southern"),
#         season = case_when(
#           hemisphere == "Northern" & adjusted_month %in% 3:5   ~ "Spring",
#           hemisphere == "Northern" & adjusted_month %in% 6:8   ~ "Summer",
#           hemisphere == "Northern" & adjusted_month %in% 9:11  ~ "Autumn",
 #          hemisphere == "Northern" & adjusted_month %in% c(12,1,2) ~ "Winter",
  #         hemisphere == "Southern" & adjusted_month %in% 3:5   ~ "Autumn",
  #         hemisphere == "Southern" & adjusted_month %in% 6:8   ~ "Winter",
  #         hemisphere == "Southern" & adjusted_month %in% 9:11  ~ "Spring",
#           hemisphere == "Southern" & adjusted_month %in% c(12,1,2) ~ "Summer",
#           TRUE ~ NA_character_
#         ))
# occ_ob2 <- occ_ob3 %>% filter(!is.na(season))
#range_path<-"F:/whaleshark_sdm/range_map/R_typus_wip2.shp"
## check sampling bias
bias_test<-bias_check(occ_data=occ_ob2,
                        temporal_level="quarter",
                        spatial_method="range_map",
                        range_path=range_path,
                        prj = "+proj=longlat +datum=WGS84")
bias_test$Plots
occ_thinned<-bias_reduce(occ_data=occ_ob2,
                           bias_result=bias_test,
                  min_dist_km = 56,
                          thin_reps = 50,
                          temporal_level = "quarter",
                          seed = 123)
bias_test<-bias_check(occ_data=occ_thinned,  
                        temporal_level="quarter",
                        spatial_method="range_map",
                        range_path=range_path,
                        prj = "+proj=longlat +datum=WGS84")
bias_test$Plots 


#### save the thinned occurrences
write.table(occ_thinned, "F:/whaleshark_sdm/occ/R_typus_qs.txt", 
   append = FALSE, sep = "\t", dec = ".",row.names = FALSE, col.names = TRUE)
occ_q1<-occ_thinned[which(occ_thinned$quarter=="Q1"),]
write.table(occ_q1, "F:/whaleshark_sdm/occ/R_typus_q1.txt", 
 append = FALSE, sep = "\t", dec = ".",row.names = FALSE, col.names = TRUE)
occ_q2<-occ_thinned[which(occ_thinned$quarter=="Q2"),]
write.table(occ_q2, "F:/whaleshark_sdm/occ/R_typus_q2.txt", 
 append = FALSE, sep = "\t", dec = ".",row.names = FALSE, col.names = TRUE)
occ_q3<-occ_thinned[which(occ_thinned$quarter=="Q3"),]
write.table(occ_q3, "F:/whaleshark_sdm/occ/R_typus_q3.txt", 
   append = FALSE, sep = "\t", dec = ".",row.names = FALSE, col.names = TRUE)
occ_q4<-occ_thinned[which(occ_thinned$quarter=="Q4"),]
write.table(occ_q4, "F:/whaleshark_sdm/occ/R_typus_q4.txt", 
   append = FALSE, sep = "\t", dec = ".", row.names = FALSE, col.names = TRUE)
```
Please note that the spatial bias still exist but have been largely reduced. A certain level of spatial autocorrelation of occurrences are ecological meaningful given species are rarely uniformly distributed in the ocean (due to the clumping nature of resources and sometimes species natural schooling or aggregation habits), and such autocorrelation is also allowed for machine-learning algorithms.

# 2. Build 2D SDMs for each temporal bin and generate related figures ####
## 2.1 SDM construction for each quarter with TB_sdm()
To predict horizontal geographic distributions within each temporal bin, we created a single function TB_sdm(), which leveraged some key functions from the packages sdm  (Naimi & Araújo, 2016), blockCV (Valavi et al., 2018), sf (Pebesma, 2018), terra (Hijmans et al., 2022), and Metrics (Hamner & Frasco, 2012). This function streamlines multiple modelling steps: (i) reading occurrences and predictors, (ii) creating folds for spatial block cross-validation, (iii) pseudo absence generation, (iv) collinearity diagnostic and selection of predictors (in an interactive manner), (v) model calibration and evaluation, (vi) ensemble construction and final prediction, and (vii) variable importance and partial dependence data extraction. It constructs models based on the sdm package, which supports classical statistical methods, machine learning algorithms, envelope methods, as well as their ensembles (see details in Naimi & Araújo, 2016). 

  Here below are the parameters used in the TB_sdm().

@param env_path character. Directory containing environmental rasters (.asc or .tif).

@param occ_path character. Path to occurrence file (tab‑delimited).

@param res_path character. Output directory.

@param species character. Column name for species (not used, kept for compatibility).

@param x character. Column name for longitude.

@param y character. Column name for latitude.

@param pre_abs_ratio numeric. Ratio of presences to pseudo-absences points.

@param collinearity_thresh numeric. Correlation threshold.

@param remove_collinear logical. Remove correlated variables.

@param max_dist numeric. Maximum distance (unit: meters) to be used in calculating Moran's I. Default is NULL

@param block_size numeric. Block size (unit: meters) can be user defined if you did not want to find it through the time-consuming auto-process based on Moran's I curve

@param algorithms character vector. sdm methods.

@param cv_method character. type of assignment of spatial blocks into folds. Can be random (default), systematic, or checkerboard.

@param n_folds integer. Number of folds for spatial cross-validation.

@param var.importance logical.

@param response.curve logical.

@param ensemble.method character. "weighted" or "unweighted".

@param ensemble.stat character. "AUC" or "TSS".

@param expr character. Select which models will be used for ensemble based on AUC and/or TSS, e.g. expr = 'auc > 0.8 & tss > 0.6'.

@param ncore integer. Number of cores.

@param seed integer. Random seed.

@param proj_path character. Optional projection rasters directory.

```{r build 2D-SDM, cache=TRUE}
#### Prior to this step, please rename the environmental predictors to make the
#### the names of variables of each quarterly bin the same as follows:
#### Depth, Mixed_layer_depth, Dissolved_oxygen, Temperature, Salinity, 
####      Net_primary_productivity, Sea_surface_height.
## Q1
rm()
gc()
# You can get access to the Q1 occurrences of whale sharks in the system file folder extdata.
d_occ <- system.file("extdata", "R_typus_q1.txt", package = "OceanSDM")
d_env <- system.file("extdata", "Q1", package = "OceanSDM")
d_prj<-system.file("extdata", "proj", package = "OceanSDM")  # Here I set up a folder where predictors of the 2nd quarter were stored, and was used as a faked future climate scenarios to demonstrate how future projections can be done.
d_res<-"F:/whaleshark_sdm/result_sdm/Q1" # Change to your own directory
#
# library(sdm)
sdm_q1<- TB_sdm(env_path=d_env, occ_path=d_occ, res_path=d_res,
                species="Species", x="x", y="y",
                algorithms = c("brt", "rf", "svm"),
                cv_method = "random", n_folds =5,
                pre_abs_ratio = 0.3, block_size = 500000,
                var.importance = TRUE, response.curve = TRUE,
                ensemble.method = "weighted", ensemble.stat = "TSS",
                ncore = 1, seed = 123,
                proj_path = d_prj,
                collinearity_thresh = 0.75, remove_collinear = T)
## Q2
d_occ <- system.file("extdata", "R_typus_q2.txt", package = "OceanSDM")
d_env <- system.file("extdata", "Q2", package = "OceanSDM")
d_res<-"F:/whaleshark_sdm/result_sdm/Q2"
sdm_q2<- TB_sdm(env_path=d_env, occ_path=d_occ, res_path=d_res,
                species="Species", x="x", y="y",
                algorithms = c("brt", "rf", "svm"),
                cv_method = "random", n_folds =5,
                pre_abs_ratio = 0.3, block_size = 500000,
                var.importance = TRUE, response.curve = TRUE,
                ensemble.method = "weighted", ensemble.stat = "TSS",
                ncore = 1, seed = 123,
                collinearity_thresh = 0.75, remove_collinear = T)
# # Q3
d_occ <- system.file("extdata", "R_typus_q3.txt", package = "OceanSDM")
d_env <- system.file("extdata", "Q3", package = "OceanSDM")
d_res<-"F:/whaleshark_sdm/result_sdm/Q3"
sdm_q3<- TB_sdm(env_path=d_env, occ_path=d_occ, res_path=d_res,
                species="Species", x="x", y="y",
                algorithms = c("brt", "rf", "svm"),
                cv_method = "random", n_folds =5,
                pre_abs_ratio = 0.3, block_size = 500000,
                var.importance = TRUE, response.curve = TRUE,
                ensemble.method = "weighted", ensemble.stat = "TSS",
                ncore = 1, seed = 123,
                collinearity_thresh = 0.75, remove_collinear = T)
# # Q4
d_occ <- system.file("extdata", "R_typus_q4.txt", package = "OceanSDM")
d_env <- system.file("extdata", "Q4", package = "OceanSDM")
d_res<-"F:/whaleshark_sdm/result_sdm/Q4"
sdm_q4<- TB_sdm(env_path=d_env, occ_path=d_occ, res_path=d_res,
                species="Species", x="x", y="y",
                algorithms = c("brt", "rf", "svm"),
                cv_method = "random", n_folds =5,
                pre_abs_ratio = 0.3, block_size = 500000,
                var.importance = TRUE, response.curve = TRUE,
                ensemble.method = "weighted", ensemble.stat = "TSS",
                ncore = 1, seed = 123,
                collinearity_thresh = 0.75, remove_collinear = T)
## save data
# save(sdm_q1,sdm_q2,sdm_q3,sdm_q4, file = "F:/whaleshark_sdm/example.RData")

```

## 2.2 Plot predictive maps, variable importance and species response curves with TB_map(), TB_varimp(), and TB_rescur() #
After successful 2D-SDM construction, TB_map(), TB_varimp(), and TB_rescur() automatically produces related figures (i.e., probability and binary maps, variable importance, and species response curves) required for publication.

  Here below are the parameters used in the TB_map().

@param root_path Character. Path to the root directory containing time period subfolders.

@param time_type Character. Type of time periods: "quarter", "month", "season".
  Default "quarter". Ignored if `time_folders` is provided.

@param time_folders Optional character vector of folder names (e.g., c("Q1","Q2","Q3","Q4")).
  If provided, overrides `time_type`.

@param time_labels Optional character vector of labels for the time periods
  (used in plot row labels). Default uses folder names.

@param output_width Numeric. Width of the output figure in inches. Default: 12.

@param output_height Numeric. Height of the output figure (per figure) in inches.
  For monthly splits, each figure will have this height. Default: 16.

@param output_filename_base Character. Base name for output TIFF files.
  For monthly data, additional suffixes are added. Default: "sdm_plot".

@param split_monthly Logical. For monthly data, if TRUE (default), split into
  three figures (Jan-Apr, May-Aug, Sep-Dec). If FALSE, produce a single figure
  with 12 rows (very tall).



  Here below are the parameters used in the TB_varimp().

@param root_path Character. Path to the root directory containing time period subfolders.

@param time_type Character. Type of time periods: "quarter", "month", "season".
  Default "quarter". Ignored if `time_folders` is provided.

@param time_folders Optional character vector of folder names (e.g., c("Q1","Q2","Q3","Q4")).
  If provided, overrides `time_type`.

@param time_names Optional character vector of labels for the time periods
  (used in legend). Default uses folder names.

@param file_name Character. Name of the variable importance CSV file within each
  subfolder (e.g., "varimp.csv"). Default: "varimp.csv".

@param method Character. Which importance metric to use: either "corTest" or
  "AUCtest". Default "AUCtest".

@param colors Optional character vector of colors, length equal to number of
  time periods. If NULL (default), uses ggplot2's default discrete color palette.

@param title Character. Plot title. Default NULL.

@param x_label Character. x-axis label. Default: "Importance".

@param y_label Character. y-axis label. Default: "Variable".

@param group_label Character. Label for the temporal bins (e.g., Quarters, Seasons, Months).
  If NULL, automatically set based on `time_type`.

@param width Numeric. Plot width in inches for saving. Default: 8.

@param height Numeric. Plot height in inches for saving. Default: 6.

@param save_path Character. Path to save the plot (TIFF format). If NULL,
  the plot is not saved.



  Here below are the parameters used in the TB_rescur().

@param root_path Character. Path to the root directory containing time period subfolders.

@param time_type Character. Type of time periods: "quarter", "month", "season".
  Ignored if `time_folders` is provided.

@param time_folders Optional character vector of folder names (e.g., c("Q1","Q2","Q3","Q4")).
  If provided, overrides `time_type`.

@param time_names Optional character vector of labels for the time periods
  (used in legend). Default uses folder names.

@param file_name Character. Name of the response curve CSV file within each
  subfolder (e.g., "rcurve.csv"). Default: "rcurve.csv".

@param colors Optional character vector of colors, length equal to number of
  time periods. Defaults to `grDevices::hcl.colors()` with palette "Set3"
  (or `rainbow` if more than 12).

@param ncol Integer. Number of columns in the panel.

@param nrow Integer. Number of rows in the panel. Automatically calculated if `NULL`.

@param y_label Character. y‑axis label. Default: `"Habitat suitability"`.

@param legend_position Character. Legend position: `"vertical"` (stacked legend
  in the last panel) or `"bottom"` (unified legend at the bottom).

@param width Numeric. Plot width in inches for saving.

@param height Numeric. Plot height in inches for saving.

@param save_path Character. Path to save the plot (TIFF format). If `NULL`,
  the plot is not saved.

```{r plot 2D-SDM, cache=TRUE}
## Clean the R environment
rm()
gc()
root_path<-system.file("extdata","result_sdm",package = "OceanSDM")
## Plot predictive maps
map_plot<-TB_map(root_path,
  time_type = "quarter",
  output_width = 8,
  output_height = 12,
  output_filename_base = "SDM_quarterly")
print(map_plot)

# Name each temporal bin
time_names <- c("1st Quarter", "2nd Quarter", "3rd Quarter", "4th Quarter")

#  Plot variable importance
imp_plot <- TB_varimp(root_path,
                      time_type = "quarter",
                      file_name = "VarImportance_Table.csv",
                      time_names = time_names,
                      method = "AUCtest",
                      colors = c("1st Quarter" = "#FFB347",
                                 "2nd Quarter" = "#4CAF50",
                                 "3rd Quarter" = "#FF6B4A",
                                 "4th Quarter" = "#5D9BEC"),
                      title = NULL,
                      x_label = "Importance",
                      y_label = NULL,
                      group_label= "Quarters",
                      width = 8,
                      height = 6,
                      save_path = "F:/whaleshark_sdm/result_sdm/variable_importance.tif")
print(imp_plot)
#  Plot response curves panel (legend at the bottom)
res_plot <- TB_rescur(root_path,
                      time_type = "quarter",
                      file_name = "rcurve_Table.csv",
                      colors = c("1st Quarter" = "#FFB347",
                                 "2nd Quarter" = "#4CAF50",
                                 "3rd Quarter" = "#FF6B4A",
                                 "4th Quarter" = "#5D9BEC"),
                      time_names = time_names,
                      time_folders = NULL,
                      legend_position = "vertical",
                      ncol=3, nrow =NULL,
                      width = 10,
                      height = 4,
                      save_path = "F:/whaleshark_sdm/result_sdm/response_curves.tif")
print(res_plot)

```

# 3. Estimate suitable/preferred depth and map the 3D habitat for each temporal bin
## 3.1 Estimate climatic niche (temperature) with niche_tag() or niche_occ() ##
A key contribution of this package is providing tools to estimate species realized climatic niche based on either tagging studies or species traits about water-column preference. I correspondingly developed two functions, niche_tag() and niche_occ(). Both functions will save a dataframe "summary_niche_fit.csv", which stored the estimated optimal temperature, preferred temperature range, and tolerable temperature range. It will be used as an input for TB_depth(). The niche fitting plot was also saved.

### 3.1.1 Use niche_tag to estimate thermal niche based on temperature utilization data from tagging studies
Many studies have indicated that the observed depths of highly-mobile marine fishes are primarily determined by physiological constraints (especially thermal niche and oxygen demand) (Arrowsmith et al., 2021; Coffey et al., 2017; Dahms & Killen, 2023; Valle et al., 2024), prey availability (Braun et al., 2023), and life history needs (e.g., reproduction, nursery, and wintering) (Andrzejaczek et al., 2019). While prey availability and life history needs are vital, these factors are currently difficult to monitor or correlate with depths. Nevertheless, species realized climatic niche (e.g., the observed thermal niche) is a result of the integrated impacts of such physical and biological factors and can be correlated with depths. The major challenge is thus how to quantify species realized climatic niche.

An optimal approach to estimate species realized climatic niche is using tag-based frequency of occupancy across in-situ ocean climate bins (e.g., temperature bins), which can be readily extracted from figures (using WebPlotDigitizer) in published tagging studies. niche_tag() estimates tolerable climate range (TCR) and preferred climate range (PCR) from such data. It first combines frequency data across studies using a weighted average approach (see the equation in the paper).

The function then allows users to choose suitable approaches to estimate TCR and PCR.

  Here below are the parameters used in the niche_tag().

@param data_list List. A list where each element is a data frame from one article,
containing columns: {bin_lower}, {bin_upper}, {percentage}, {article_weight}.

@param global_bins Data frame. A data frame with columns: {bin_lower}, {bin_upper} defining standardized bins.

@param samples_per_article Numeric. Number of simulated samples per article for
standardization (default: 10000).

@param pref_prob Numeric. Proportion of area on each side of the mode to include in the
preference range. Default is 0.5 (50% of each side's area).

@param toler_prob Numeric. Proportion of area on each side of the mode to include in the
tolerance range. Default is 0.95 (95% of each side's area).

@param bw_method Character. Bandwidth selection method for kernel density estimation.
Options are:
  {"nrd0"}: Silverman's rule-of-thumb (default in {stats::density}).
        Fast and stable, suitable for unimodal, near-normal distributions, but may
        oversmooth multimodal or skewed distributions.
  {"nrd"}: Silverman's rule-of-thumb (variant). Slightly different
        scaling factor, similar properties to {"nrd0"}.
  {"SJ"}: Sheather–Jones direct plug-in estimator. Adaptive to
        data structure, preferred for complex/multimodal distributions, but
        computationally more intensive.
  {"bcv"}: Biased cross-validation. Tends to produce smaller bandwidths,
        useful for fine-scale features, but can be unstable.
  {"ucv"}: Unbiased cross-validation. Tends to produce larger bandwidths,
        more stable than {"bcv"} but computationally intensive.
        
@param variable_name Character. Name of the variable (e.g., {"Temperature"},
{"Dissolved Oxygen"}) for plotting (default: {"Environmental variable"}).

@param unit Character. Unit of the variable (e.g., {"°C"}, {"mg/L"}) for plotting

@param seed Integer. Random seed for reproducibility (default: 123).

@param width Numeric. Plot width in inches (default: 6).

@param height Numeric. Plot height in inches (default: 5).

@param save_path Character. Directory path to save plot and summary CSV.

```{r niche_tag,cache=TRUE}
## Article 1 data (used as the standard bin given the no. of tagged individuals is the highest)
article1_df <- data.frame(
  bin_lower = c(0, 3, 6, 9, 12, 15, 18, 21, 24, 27, 30, 33, 36, 39),
  bin_upper = c(3, 6, 9, 12, 15, 18, 21, 24, 27, 30, 33, 36, 39, 42),
  percentage = c(0,0,0,0,0,0,0.3, 11, 29, 37, 16, 6, 0.4, 0.3) / 100,
  article_weight = 50  # Sample size: 50 individuals
)

## Article 2 data (different binning scheme)
article2_df <- data.frame(
  bin_lower = c(10, 12.5, 15, 17.5, 20, 22.5, 25, 27.5, 30),
  bin_upper = c(12.5, 15, 17.5, 20, 22.5, 25, 27.5, 30, 32.5),
  percentage =c(0.8856, 0.8856, 0.6642, 0.4428, 0.4428, 4.6494, 29.4465, 59.3358, 2.8782)/100 ,
  article_weight = 8  # Sample size: 8 individuals
)

## Article 3 data (another different binning scheme)
article3_df <- data.frame(
  bin_lower = c(10, 12, 14, 16, 18, 20, 22, 24, 26, 28),
  bin_upper = c(12, 14, 16, 18, 20, 22, 24, 26, 28, 30),
  percentage = c(0.65445, 3.02054, 4.36468, 4.36468, 4.36468, 6.207209, 12.58558, 25.50846, 26.01188, 12.91784)/100 ,
  article_weight = 1  # Sample size: 1 individuals
)

## Article 4 data (another different binning scheme)
article4_df <- data.frame(
  bin_lower = c(0, 2, 5, 10, 15, 17.5, 20, 22, 25, 27, 29,31),
  bin_upper = c(2, 5, 10, 15, 17.5, 20, 22, 25, 27, 29,31,33),
  percentage =  c(0.039048, 0.002092, 0.055783, 2.545097, 4.554678, 8.029258, 18.46834, 44.00787, 12.77708, 8.686102, 0.594088, 0.240564)/100,
  article_weight = 15  # Sample size: 15 individuals
)

## Article 5 data (another different binning scheme)
article5_df <- data.frame(
  bin_lower = c(0, 5, 10, 15, 18, 20, 22, 24, 26, 28, 30, 32, 34),
  bin_upper = c(5, 10, 15, 18, 20, 22, 24, 26, 28, 30, 32, 34,36),
  percentage =  c(0.002896, 0.004234, 0.013009, 0.012676, 0.011109, 0.018206, 0.055804, 0.443287, 0.41829, 0.020489,0,0,0),
  article_weight = 27  # Sample size: 15 individuals
)
## Article 6 data (another different binning scheme)
article6_df <- data.frame(
  bin_lower = c(4,8,12,16,20,22,24,26,28),
  bin_upper = c(8,12,16,20,22,24,26,28,30),
  percentage =  c(0.0188, 0.0075, 0.0063, 0.0088, 0.0151, 0.1217, 0.3111, 0.4027, 0.1116),
  article_weight = 17  # Sample size: 15 individuals
)
## Article 7 data (another different binning scheme)
article7_df <- data.frame(
  bin_lower = c(10,12.5,15,17.5,20,22.5,25,27.5,30,32.5),
  bin_upper = c(12.5,15,17.5,20,22.5,25,27.5,30,32.5,35),
  percentage =  c(0, 0.029412, 0.011765, 0.011765, 0.019608, 0.060785, 0.372549, 0.37647, 0.115686, 0.001961),
  article_weight = 17  # Sample size: 15 individuals
)

## Create list of articles
data_list <- list(article1_df, article2_df, article3_df,article4_df,article5_df,article6_df,article7_df)

## Define unified global binning scheme (follow article1_df)
global_bins <- article1_df[,1:2]
## Run estimation
temp_fit<-niche_tag(
    data_list,
    global_bins,
    toler_prob = 0.99,
    pref_prob = 0.5,
    bw_method = "nrd0",
    variable_name = "Temperature",
    unit = "℃",
    seed = 123,
    width = 6,
    height = 5,
    save_path = "F:/whaleshark_sdm/temp_fit/tag"
)

```


### 3.1.2 Extract temperature data from thinned occurrences to estimate climatic niche
When tag-based temperature data are unavailable, niche_occ() implements an alternative, trait based approach. It utilizes bias-reduced occurrences to extract real-time temperatures from the depth layer preferred by species (e.g., surface for pelagic fish) rather than depths recorded in biodiversity databases as in the voluModel package. It then allows users to select similar methods (KDE-based probability) to estimate TCR and PCR as mentioned for niche_tag().


  Here below are the parameters used in the niche_occ().

@param occ_path character. Path to occurrence file (table with columns: long, lat, year, month).

@param env_path character. Path to NetCDF file containing environmental data (e.g., temperature, DO).

@param pref_prob numeric. Proportion of area on each side of the mode to include in the
preference range. Default is 0.5 (50% of each side's area).

@param toler_prob numeric. Proportion of area on each side of the mode to include in the
tolerance range. Default is 0.95 (95% of each side's area).

@param bw_method Character. Bandwidth selection method for kernel density estimation.
Options are:
  {"SJ"} (default): Sheather–Jones direct plug-in estimator. Adaptive to
        data structure, preferred for complex/multimodal distributions, but
        computationally more intensive.
  {"nrd0"}: Silverman's rule-of-thumb (default in {stats::density}).
        Fast and stable, suitable for unimodal, near-normal distributions, but may
        oversmooth multimodal or skewed distributions.
  {"nrd"}: Silverman's rule-of-thumb (variant). Slightly different
        scaling factor, similar properties to {"nrd0"}.
  {"bcv"}: Biased cross-validation. Tends to produce smaller bandwidths,
        useful for fine-scale features, but can be unstable.
  {"ucv"}: Unbiased cross-validation. Tends to produce larger bandwidths,
        more stable than {"bcv"} but computationally intensive.
        
@param save_path character. Directory to save output plot and summary CSV. If NULL, nothing saved.

@param min_year numeric. First year of occurrence data (used to compute time index in NetCDF).

@param var_name character. Variable name in the NetCDF file (e.g., "thetao_mean", "o2").

@param env_label character. Display name for the environmental variable (e.g., "Temperature", "Dissolved Oxygen"). Defaults to var_name.

@param unit character. Unit of the environmental variable (e.g., "°C", "mg/L"). Defaults to empty string.

@param lon_range numeric vector of length 2: min and max longitude of the raster extent.

@param lat_range numeric vector of length 2: min and max latitude of the raster extent.

@param width numeric. Plot width in inches (default: 6).

@param height numeric. Plot height in inches (default: 5).

@param seed integer. Random seed for reproducibility (default: 123).

Here, based on the pelagic trait of the focal species, I used niche_occ() to extract the temperature at the depth of 1 m at the thinned occurrences to estimate thermal niche.
```{r niche_occ,cache=TRUE}
occ_path<-system.file("extdata", "R_typus_qs.txt", package = "OceanSDM")
# ocean temperature at the surface layer (depth = 1m)
temp_path<-system.file("extdata", "cms/IndoWPac_temp_1m_y2015_y2024.nc", package = "OceanSDM")
save_path <- "F:/whaleshark_sdm/temp_fit/habitat"
thermal_occ<-niche_occ(
  occ_path,
  temp_path,
  pref_prob = 0.5,
  toler_prob = 0.99,
  bw_method = "nrd0",
  save_path,
  min_year = 2015,
  var_name = "thetao_mean",
  env_label = "Temperature",
  unit = "℃",
  lon_range=c(90,180),
  lat_range=c(-41,40),
  width = 6,
  height = 5
  )
```

### 3.1.3 Extract oxygen data (optional)
Although the depth limits might be best estimated with thermal niche, other physiological constraints such as demand of oxygen can be estimated as well with data in a user-defined depth at species occurrences. 
```{r oxygen niche,cache=TRUE}
# occ_path<-system.file("extdata", "R_typus_qs.txt", package = "OceanSDM")
# var_name<-"o2"
## Download the O2 data with env_download() first
# o2_path<-"F:/cmems/IndoWPac_o2_1m_y2015_y2024.nc" # dissolved oxygen at the surface layer (depth = 1m)
# save_path <- "F:/whaleshark_sdm/o2_fit/habitat"
# 
# oxy_occ<-niche_occ(
#   occ_path,
#   o2_path,
#   pref_prob = 0.5,
#  toler_prob = 0.99,
#  bw_method = "nrd0",
#   save_path,
#   min_year=2015,
#   var_name = "o2",
#   env_label = "Dissolved oxygen",
#   unit = "μmol/L",
#   lon_range=c(90,180),
#  lat_range=c(-41,40),
#   width = 8,
#   height = 5
# )
```

## 3.2 Create temporally-binned temperature for each depth layer using clim_bin() ##
Before predicting vertical distributions across space and time, clim_bin() calculates the average climate value at each depth layer of each predicted suitable pixel in each temporal bin (derived from TB_sdm).

  Here below are the parameters used in the niche_occ().

@param clim_nc Character. Path to the multi-depth layers over years
  and months of climate NetCDF file.

@param presence_asc List or SpatRaster. A named list where each element is a
  presence raster (values 1 = presence, NA/0 = absence) corresponding to a
  temporal group (folder name). For season, names must be one of
  "Winter", "Spring", "Summer", "Autumn".

@param time_type Character. Type of temporal aggregation. One of `"month"`,
  `"quarter"`, or `"season"`.

@param var_name Character. Variable name in NetCDF. If not found exactly,
  the function will attempt to match (case‑insensitive) and warn.
  e.g., `"thetao_mean"`.

@param save_path Character. Directory path where output files will be saved.

@param na_value Numeric. Value to replace `NA`. Default `-9999`.

```{r clim_bin,cache=FALSE}
## Collect predicted binary maps of each temporal bin (here, quarters)
quarter_files<-c(
  Q1=system.file("extdata/result_sdm/Q1/predicted_presence.asc",package = "OceanSDM"),
  Q2=system.file("extdata/result_sdm/Q2/predicted_presence.asc",package = "OceanSDM"),
  Q3=system.file("extdata/result_sdm/Q3/predicted_presence.asc",package = "OceanSDM"),
  Q4=system.file("extdata/result_sdm/Q4/predicted_presence.asc",package = "OceanSDM"))
presence_asc <- lapply(quarter_files, terra::rast)
save_path<-"F:/whaleshark_sdm/4D"

## In subsection 1.2.2, you can run the code to downloaded temperature data for three water columns: 0-200m, 200-400m, 400-1050m. And then you can employ clim_bin() to create temporally-binned predictors for each depth layer based on these three .nc files.
# For the first water column (0-200m)
temp_nc<-"F:/whaleshark_sdm/cmems/IndoWPac_temp_200m_y2015_y2024.nc"
clim_bin(temp_nc, presence_asc, time_type = "quarter",
         save_path=save_path, na_value = -9999)
## You can run the following code for the other two water columns by removing the hashtags.
# temp_nc<-"F:/whaleshark_sdm/cmems/IndoWPac_temp_200_400m_y2015_y2024.nc"
# clim_bin(temp_nc, presence_asc, time_type = "quarter",
#          save_path=save_path,  na_value = -9999)
# temp_nc<-"F:/whaleshark_sdm/cmems/IndoWPac_temp_400_1050m_y2015_y2024.nc"
# clim_bin(temp_nc, presence_asc, time_type = "quarter",
#          save_path=save_path, na_value = -9999)
```

## 3.3 Calculate TD and PD for each temporal bin with TB_depth() ######
TB_depth() calculates tolerable depths (TD) and preferred depths (PD) in each pixel for each temporal bin by comparing TCR and PCR, respectively, with the climate value along the vertical gradient. It outputs three statistics (minimum, maximum, range) for TD and PD and plots the vertical climate profile for a random sample (default, 100) of grids. However, this process is very time-consuming (lasting > 8 hrs) given the function has to calculate these statistics grid by grid and there are 47 depth layers. Therefore, I did not run the following code when generating this vignette. But I have stored the outputs while producing the paper.

  Here below are the parameters used in the TB_depth().

@param temp_bin_path Character. Root path of temp_bin generated files.

@param PTR Numeric vector of length 2. Temperature range, e.g., `c(23.5, 30.2)`.

@param time_type Character. Time type, should match that used in temp_bin:
  `"month"`, `"quarter"`, or `"season"`.

@param layer_option Character. Which depth interval to extract: `"top"` (shallowest)
  or `"bottom"` (deepest). Default is `"top"`.

@param save_path Character. Path to save output files.

@param na_value Numeric. Value to replace NA (default: -9999).

@param use_makima Logical. Whether to use makima interpolation. If `FALSE` , falls back to spline.

@param N_cells Integer. Number of random cells to plot (5--30). If `0`, no plot is generated.

@param ncores Integer. Number of cores for parallel processing (uses `pbapply`). Default: 1.

```{r TB_depth(),cache=TRUE}
##### Example shown with estimates based on tag data
df<-read.csv(system.file("extdata","summary_niche_fit.csv",package = "OceanSDM"))
PTR<-df[2:3,2] # preferred temperature range
TTR<-df[4:5,2] # tolerable temperature range
PTR
TTR
## Run the whole code in subsection 3.2 to create temporally-binned mean temperature for each depth layer. Here I stored all these files in "F:/whaleshark_sdm/4D"
temp_bin_path<-system.file("extdata","4D",package = "OceanSDM") # Change to your own directory
save_path_TPD<-"F:/whaleshark_sdm/4D/TPD" ## for storing data of Preferred Depth (PD)
save_path_TTD<-"F:/whaleshark_sdm/4D/TTD" ## for storing data of Tolerable Depth (TD)
## For Tolerable Depth
TTD_quarter<-TB_depth(temp_bin_path, TTR, time_type = "quarter",
                      layer_option="top",
                      save_path_TTD, na_value = -9999,
                      use_makima = TRUE,
                      N_cells = 50)
## For Preferred Depth (PD), run the following code by removing the hashtags:
# TPD_quarter<-TB_depth(temp_bin_path, PTR, time_type = "quarter",
#                       layer_option="top",
#                       save_path_TPD, na_value = -9999,
#                       use_makima = TRUE,
#                       N_cells = 50)
```

## 3.4 Plot habitat volume and area with plot_stat_4D() ##############
plot_stat_4D() maps the statistics (Max, Min, Range) of TD and PD across predicted 2D presence maps, generates boxplots, and tests for significant differences among temporal bins using Kruskal Wallis followed by Dunn’s post-hoc tests. It also create a figure to showcase the total suitable habitat volume based on the TD or PD, and estimates total suitable habitat area based on the 2D presence map of each temporal bin. 

  Here below are the parameters used in the plot_stat_4D().

@param file_list Character vector. File paths to raster files.

@param save_path Character. Directory path to save plots and report.

@param time_type Character. One of {"quarter"}, {"month"}, or {"season"}.

@param presence_files Character vector (optional). Paths to binary presence rasters.

@param plot_overlay Logical. Whether to create volume-area overlay plot.

@param sample_fraction Numeric. Fraction of complete cells to sample for statistical tests. Default: 0.2 (20\%).


```{r plot_stat_4D,cache=FALSE}
## Run the whole code in subsection 3.3 to estimate three parameters (minimum, maximum, range) of thermally preferred depth or thermally tolerable depth. Here I load the input data I already prepared.

TPD_quarter<-list(
system.file("extdata","4D/TPD/Q1/Q1_min_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TPD/Q1/Q1_max_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TPD/Q1/Q1_depth_range.asc",package="OceanSDM"),
  system.file("extdata","4D/TPD/Q2/Q2_min_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TPD/Q2/Q2_max_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TPD/Q2/Q2_depth_range.asc",package="OceanSDM"),
  system.file("extdata","4D/TPD/Q3/Q3_min_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TPD/Q3/Q3_max_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TPD/Q3/Q3_depth_range.asc",package="OceanSDM"),
  system.file("extdata","4D/TPD/Q4/Q4_min_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TPD/Q4/Q4_max_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TPD/Q4/Q4_depth_range.asc",package="OceanSDM")
)
TTD_quarter<-list(
  system.file("extdata","4D/TTD/Q1/Q1_min_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TTD/Q1/Q1_max_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TTD/Q1/Q1_depth_range.asc",package="OceanSDM"),
  system.file("extdata","4D/TTD/Q2/Q2_min_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TTD/Q2/Q2_max_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TTD/Q2/Q2_depth_range.asc",package="OceanSDM"),
  system.file("extdata","4D/TTD/Q3/Q3_min_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TTD/Q3/Q3_max_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TTD/Q3/Q3_depth_range.asc",package="OceanSDM"),
  system.file("extdata","4D/TTD/Q4/Q4_min_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TTD/Q4/Q4_max_depth.asc",package="OceanSDM"),
  system.file("extdata","4D/TTD/Q4/Q4_depth_range.asc",package="OceanSDM")
)
save_path_TPD<-"F:/whaleshark_sdm/4D/TPD" ## for Preferred Depth (PD)
save_path_TTD<-"F:/whaleshark_sdm/4D/TTD" ## for Tolerable Depth (TD)
quarter_files<-c(
  Q1=system.file("extdata","result_sdm/Q1/predicted_presence.asc",package="OceanSDM"),
  Q2=system.file("extdata","result_sdm/Q2/predicted_presence.asc",package="OceanSDM"),
  Q3=system.file("extdata","result_sdm/Q3/predicted_presence.asc",package="OceanSDM"),
  Q4=system.file("extdata","result_sdm/Q4/predicted_presence.asc",package="OceanSDM"))

result_TPD <- plot_stat_4D(
  file_list = TPD_quarter,
  save_path = save_path_TPD,
  time_type = "quarter",
  presence_files = quarter_files
)
result_TPD$depth_range
result_TPD$depth_min
result_TPD$depth_max
result_TPD$overlay_plot
result_TTD <- plot_stat_4D(
  file_list = TTD_quarter,
  save_path = save_path_TTD,
  time_type = "quarter",
  presence_files = quarter_files
)
result_TTD$depth_range
result_TTD$depth_min
result_TTD$depth_max
result_TTD$overlay_plot
volum<-rbind(result_TPD$volum,result_TTD$volum)
volum$Group<-c(rep("PD",4),rep("TD",4))
area<-result_TPD$area
write.csv(volum,"F:/whaleshark_sdm/4D/volume_data.csv",row.names = F)
write.csv(area,"F:/whaleshark_sdm/4D/area_data.csv",row.names = F)
```

## 3.5 Plot habitat volumes (estimated based on TD and PD) and habitat area in one graph (Optional)

You can use the following codes to plot the two habitat volume estimates (based on TD and PD) and
habitat area on one graph.
```{r plot_volume_area,cache=FALSE}
library(ggplot2)
volum<-read.csv(system.file("extdata","4D/volume_data.csv",package="OceanSDM"))
area<-read.csv(system.file("extdata","4D/area_data.csv",package="OceanSDM"))

# convert raw values to [0,1]
choose_factor <- function(x) {
if (all(x == 0)) return(1)
max_val <- max(x, na.rm = TRUE)
e <- floor(log10(max_val))
factor <- 10^e
while (max_val / factor > 10) factor <- factor * 10
while (max_val / factor < 1 && factor > 1) factor <- factor / 10
return(factor)
}
vol_factor <- choose_factor(volum$Volume_km3)
area_factor <- choose_factor(area$Area_km2)
volum$Volume <- volum$Volume_km3 / vol_factor
area$Area <- area$Area_km2 / area_factor
scale_ratio <- max(volum$Volume ) / max(area$Area)
area$Area <- area$Area * scale_ratio
# Prepare plot labels
vol_exp <- log10(vol_factor)
area_exp <- log10(area_factor)
vol_label <- bquote("Habitat volume (×" ~ 10^.(vol_exp) ~ " km"^3 ~ ")")
area_label <- bquote("Area (×" ~ 10^.(area_exp) ~ " km"^2 ~ ")")
# Ensure Temporal_bin is an ordered factor (in the order Q1, Q2, Q3, Q4)
volum$Temporal_bin <- factor(volum$Temporal_bin, levels = c("Q1", "Q2", "Q3", "Q4"))
area$Temporal_bin <- factor(area$Temporal_bin, levels = c("Q1", "Q2", "Q3", "Q4"))
# Merge area data into the main dataset (for plotting)
library(dplyr)
area_matched <- area %>%
  right_join(volum %>% dplyr::select(Temporal_bin) %>% distinct(), by = "Temporal_bin")
# Calculate the range of Volume (for automatic scaling of Area)
vol_range <- range(volum$Volume, na.rm = TRUE)
area_range <- range(area_matched$Area, na.rm = TRUE)
# Scaling Factor: The numerical range mapping Area to Volume
# Mapping formula: area_scaled = (Area - area_min) / (area_max - area_min) * (vol_max - vol_min) + vol_min
area_min <- area_range[1]
area_max <- area_range[2]
vol_min <- vol_range[1]
vol_max <- vol_range[2]
# Avoid division by zero (if the area range is too small)
if (area_max == area_min) {
scaling_factor <- 1
offset <- 0
} else {
scaling_factor <- (vol_max - vol_min) / (area_max - area_min)
offset <- vol_min - scaling_factor * area_min
}
# Add a column for the scaled area
area_matched$Area_scaled <- area_matched$Area * scaling_factor + offset
p <- ggplot(volum, aes(x = Temporal_bin, y = Volume, fill = Group)) +
  geom_col(position = position_dodge(0.9), width = 0.7) +
  geom_line(
  data = area_matched,
  aes(x = Temporal_bin, y = Area_scaled, group = 1),
  color = "red", linewidth = 1,
  inherit.aes = FALSE
  ) +
  geom_point(
  data = area_matched,
  aes(x = Temporal_bin, y = Area_scaled),
  color = "red", size = 3,
  inherit.aes = FALSE
  ) +
  scale_y_continuous(
  name = vol_label,
  sec.axis = ggplot2::sec_axis(
  ~ . / scale_ratio,
  name = area_label,
  labels = function(x) format(x, digits = 2)
  )
  )+
  labs(x = "Quarters", fill = "Group") +
  theme_minimal() +
  theme(
  axis.line = element_line(),
  axis.title.y.right = element_text(color = "red"),
  axis.text.y.right = element_text(color = "red"),
  legend.position = c(0.93, 0.96),
  legend.justification = c(1, 1)
  )
p
## save the figure
ggsave(
filename = "F:/whaleshark_sdm/figures/Fig. 7 TB_volume_area.tif",
plot = p,
device = "tiff",
dpi = 300,
compression = "lzw",
width = 6,
height = 6
)
```

#################### The end ############################

# References

Andrzejaczek, S., Gleiss, A. C., Pattiaratchi, C. B., & Meekan, M. G. (2019). Patterns and drivers of vertical movements of the large fishes of the epipelagic. Reviews in Fish Biology and Fisheries, 29(2), 335–354. https://doi.org/10.1007/s11160-019-09555-1

Arrowsmith, L., Sequeira, A., Pattiaratchi, C., & Meekan, M. (2021). Water temperature is a key driver of horizontal and vertical movements of an ocean giant, the whale shark Rhincodon typus. Marine Ecology Progress Series, 679, 101–114. https://doi.org/10.3354/meps13899

Braun, C. D., Della Penna, A., Arostegui, M. C., Afonso, P., Berumen, M. L., Block, B. A., Brown, C. A., Fontes, J., Furtado, M., Gallagher, A. J., Gaube, P., Golet, W. J., Kneebone, J., Macena, B. C. L., Mucientes, G., Orbesen, E. S., Queiroz, N., Shea, B. D., Schratwieser, J., … Thorrold, S. R. (2023). Linking vertical movements of large pelagic predators with distribution patterns of biomass in the open ocean. Proceedings of the National Academy of Sciences, 120(47), e2306357120. https://doi.org/10.1073/pnas.2306357120

Coffey, D. M., Carlisle, A. B., Hazen, E. L., & Block, B. A. (2017). Oceanographic drivers of the vertical distribution of a highly migratory, endothermic shark. Scientific Reports, 7(1), 10434. https://doi.org/10.1038/s41598-017-11059-6

Dahms, C., & Killen, S. S. (2023). Temperature change effects on marine fish range shifts: A meta-analysis of ecological and methodological predictors. Global Change Biology, 29(16), 4459–4479. https://doi.org/10.1111/gcb.16770

Dobson, R., Challinor, A. J., Cheke, R. A., Jennings, S., Willis, S. G., & Dallimer, M. (2023). dynamicSDM: An R package for species geographical distribution and abundance modelling at high spatiotemporal resolution. Methods in Ecology and Evolution, 14(5), 1190–1199. https://doi.org/10.1111/2041-210X.14101

Hertz, P. E., Huey, R. B., & Stevenson. (1993). Evaluating Temperature Regulation by Field-Active Ectotherms: The Fallacy of the Inappropriate Question | The American Naturalist: Vol 142, No 5. The American Naturalist, 142(5). https://www.journals.uchicago.edu/doi/10.1086/285573

Naimi, B., & Araújo, M. B. (2016). sdm: A reproducible and extensible R platform for species distribution modelling. Ecography, 39(4), 368–375. https://doi.org/10.1111/ecog.01881

Owens, H. L., & Rahbek, C. (2023). voluModel: Modelling species distributions in three-dimensional space. Methods in Ecology and Evolution, 14(3), 841–847. https://doi.org/10.1111/2041-210X.14064

Valle, M., Ramírez-Romero, E., Ibaibarriaga, L., Citores, L., Fernandes-Salvador, J. A., & Chust, G. (2024). Pan-Atlantic 3D distribution model incorporating water column for commercial fish. Ecological Modelling, 490, 110632. https://doi.org/10.1016/j.ecolmodel.2024.110632
