---
title: "Introduction to OceanSDM"
date: 2026-06-24
output:
  html_document:
    toc: true
    toc_depth: 3
    css: vignette.css
vignette: >
  %\VignetteIndexEntry{Introduction to OceanSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Predicting the dynamic, three-dimensional distributions of marine species is challenging given most SDM tools currently can only predict static 2D habitats (but see an example of voluModel package,Owens & Rahbek, 2023). Here, we present a novel package 'OceanSDM' to fill this gap. This package includes 18 user-friendly functions helping researchers to complete tasks from initial data extraction to final publication-quality figure creation. It can be split into three major sections: (1) Prepare species and environmental data, (2) Build 2D SDMs for each temporal bin and generate related figures, and (3) Estimate suitable/preferred depth and map the 3D habitat for each temporal bin. Here I illustrate the utility of this package with a pelagic shark (whale shark). For the temporal bin, I used quarterly bins (four bins).

Here are the packages you need.
```{r setup}
library(OceanSDM) #
```
# 1. Prepare species and environmental data 
## 1.1 Download species data from GBIF and OBIS using ob_data() 
The Global Biodiversity Information Faculty (GBIF) and Ocean Biodiversity Information System (OBIS) are the two major sources of species occurrences at the global scale. The function ob_data() helps users automatically download the geo-referenced occurrences of either a single or many species. Here I demonstrated with an example of smaller temporal and spatial ranges than the real case in the paper to save the running time for creating this vignette.

  Here is a detailed explanation about this function:

-- @param key Character. The species scientific name (e.g., "Rhincodon typus").

-- @param sp_geometry Character. The geographic range of the study area as a WKT polygon
--   (e.g., "POLYGON ((20 -41, 20 40, 180 40, 180 -41, 20 -41))").

-- @param start_date Character. Start date used in `robis::occurrence()`, format "YYYY-MM-DD".

-- @param end_date Character. End date used in `robis::occurrence()`, format "YYYY-MM-DD".

-- @param event_date Character. Year range used in `rgbif::occ_search()`, format "YYYY-MM-DD,YYYY-MM-DD".

-- @return A list containing:
--   {sp_occ}{Data frame of cleaned occurrence records with columns:
     `Species`, `decimalLatitude`, `decimalLongitude`, `date`, `year`, `month`, `day`.}
--   {plot}{A ggplot map plot of occurrences.}
```{r occurrence, dev='png', cache=TRUE}
 ws_occ<-ob_data(
    key<-'Rhincodon typus' ,# It can be a vector of the Latin names of all focal species
    sp_geometry <- "POLYGON ((90 -41, 90 40, 180 40, 180 -41, 90 -41))", # Replace this to "POLYGON ((90 -41, 90 40, 180 40, 180 -41, 90 -41))" to simulate my real case study!!!
    start_date<-"2015-01-01", # Replace this to "2015-01-01" to simulate my real case study!!!
    end_date<-"2024-12-30",
    event_date<-'2015-01-01,2024-12-30') # Replace this to '2015-01-01,2024-12-30' to simulate my real case study!!!
# # save the occurrences
write.csv(ws_occ$sp_occ,"F:/whaleshark_sdm/occ/R_typus.csv",row.names = F)
```

## 1.2 Download .nc files of marine environmental predictors using env_download() 
The Copernicus Marine Service (CMS) offers dynamic, three-dimensional environmental data that can be extracted with the function env_download(). But please note you must first install the copernicusmarine toolbox (and configure its running environment with conda or other softwares, see details at https://help.marine.copernicus.eu/en/articles/7970514-copernicus-marine-toolbox-installation) on which this function was built. Here I only showed a case of extracting sea surface temperature at smaller temporal and geographic ranges than the real case I used in the paper.
  Here is a detailed explanation about this function:

-- @param path_cmems_tool Character. Path to the copernicusmarine executable,
   e.g. `"/miniconda/envs/copernicusmarine/Scripts/copernicusmarine.exe"`.

-- @param out_dir Character. Directory where output data will be stored.

-- @param productId Character. ID of the CMEMS product (see dataset user manual
   in the CMEMS product webpage). e.g., `"cmems_mod_glo_phy-mnstd_my_0.25deg_P1M-m"`.
   
-- @param variable Character. Variable name to download, prefixed with "--variable ".
    e.g., " --variable thetao_mean".
    
-- @param date_min Date or character. Minimum date. e.g., is `1993-01-01`.

-- @param date_max Date or character. Maximum date. e.g., is `2022-12-01`.

-- @param lon Numeric vector of length 2. Longitude range (min, max). e.g., `c(-180, 179.75)`.
   
-- @param lat Numeric vector of length 2. Latitude range (min, max). e.g., `c(-80, 90)`.
   
-- @param depth List of length 2. Depth range in meters (min, max). e.g., `list(0.51, 199.79)`.

-- @param out_name Character. Output file name ending with `.nc`.

-- @return No return value. The function downloads a NetCDF file to `out_dir`.
```{r set up tool path, cache=TRUE}
## Set up the path of the cms downloading tool and output directory
path_cms<-"C:/shark/miniconda/envs/copernicusmarine/Scripts/copernicusmarine.exe" # replace to your own directory
out_dir <- "F:/whaleshark_sdm/cmems" # replace to your own directory
library(lubridate)
```
### 1.2.1 Extract surface layer of marine environmental predictors
```{r download surface predictor data, cache=TRUE}
## Note here is a simplified example for fast creating the vignettes
productId <- "cmems_mod_glo_phy-mnstd_my_0.25deg_P1M-m"
variable <- " --variable thetao_mean"
date_min <- ymd(20240101) # start_date, it should be ymd(20150101) in the paper
date_max <- ymd(20241201) # end_date
lon<-c(90,180) # longitudinal range
lat<-c(-41,40) # latitudinal range
## sea surface temperature
depth <- list(0.51, 2) # depth_min, depth_max
out_name <- paste("IndoWPac_temp_1m_y2024.nc", sep="") # it should be paste("IndoWPac_temp_1m_y2015_y2024.nc", sep="") in the paper
env_download(path_cms,out_dir,productId,variable,
              date_min,date_max,lon,lat,depth,out_name)
## sea surface height
# variable <- " --variable zos_mean"
# out_name <- paste("IndoWPac_slh_1m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
## mixed layer depth
# variable <- " --variable mlotst_mean"
# out_name <- paste("IndoWPac_mlt_1m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
## salinity
# variable <- " --variable so_mean"
# out_name <- paste("IndoWPac_so_1m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
```

### 1.2.2 Extract multiple depth layers of ocean temperature
This step is NOT necessary for the 2D SDM construction, but is required for estimating suitable/tolerable depths based on species realized climatic niche and the three-dimensional distributions of the focal climatic variable (here, temperature) over the chosen temporal bins (here, four quarters). Here given the size of the multi-depth layers of data, I split it into four .nc files (0-200m, 200-400m, 400-1050m, 1050-1946m). The maximum depth (1946m) was based on the known depth records of the whale shark in the study area. The codes were set to 'not run' with the hash tag while creating this vignette as it was time consuming. Users can simply remove these hash tags (Using 'Control+Shift+C' for Windows) and run the code.
```{r download multi-depth layers of temperature, cache=TRUE}
# variable <- " --variable thetao_mean"
# path_cmems_tool="C:/shark/miniconda/envs/copernicusmarine/Scripts/copernicusmarine.exe"
# out_dir = "F:/whaleshark_sdm/cmems"
# productId = "cmems_mod_glo_phy-mnstd_my_0.25deg_P1M-m"
# date_min = ymd(20150101) # start_date
# date_max = ymd(20241201) # end_date
# lon<-c(90,180)
# lat<-c(-41,40)
# depth <- list(0.50, 200) # depth_min, depth_max
# out_name <- paste("IndoWPac_temp_200m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
# depth <- list(200, 400) # depth_min, depth_max
# out_name <- paste("IndoWPac_temp_200_400m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
# depth <- list(400, 1050) # depth_min, depth_max
# out_name <- paste("IndoWPac_temp_400_1050m_y2015_y2024.nc", sep="")
#env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
# depth <- list(1050, 1946) # depth_min, depth_max
# out_name <- paste("IndoWPac_temp_1050_1946m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
```

### 1.2.3 Extract sea surface data for net primary productivity and O2 from another CMS dataset
```{r download npp and o2, cache=TRUE}
# out_dir <- "F:/cmems"
# productId <- "cmems_mod_glo_bgc_my_0.25deg_P1M-m"
# date_min <- ymd(20150101) # start_date
# date_max <- ymd(20241201) # end_date
# lon<-c(90,180)
# lat<-c(-41,40)
# depth <- list(0.51, 2) # depth_min, depth_max
# variable <- " --variable nppv"
# out_name <- paste("IndoWPac_npp_1m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)

# variable <- " --variable o2"
# out_name <- paste("IndoWPac_o2_1m_y2015_y2024.nc", sep="")
# env_download(path_cms,out_dir,productId,variable,
#             date_min,date_max,lon,lat,depth,out_name)
```
For extracting predictor datasets of future projections, please check https://esgf-node.llnl.gov/search/cmip6/
There is also an R package for downloading CMIP6 data: https://github.com/marine-ecologist/CMIP6/blob/main/CMIP6_projections.R

### 1.2.4 Prepare temporally-binned predictors using env_bin()
To predict the dynamic distributions of marine species across the 2D space, all predictors should be aggregated into mean values within the specific temporal bins.  
```{r, cache=TRUE}
gc() # clear the R environment 
## sea surface temperature (Note this is a simplified example of using downloaded data in 1.2.1)
env_bin(sp_trait="pelagic",            
        env_path="F:/whaleshark_sdm/cmems/IndoWPac_temp_1m_y2024.nc", 
# in the paper, I used "F:/whaleshark_sdm/cmems/IndoWPac_temp_1m_y2015_y2024.nc"
       out_dir="F:/whaleshark_sdm/bin2",  
# in the paper, I used "F:/whaleshark_sdm/bin". here I set the directory to a temporal folder to receive the output from this example.
        var_name="thetao_mean",
        lon_min=90,
       lon_max=180,
        lat_min=-41,
        lat_max=40, # in the paper, I used 40
       date_min=ymd(20240101), # in the paper, I used ymd(20150101)
        date_max = ymd(20241201),
        temporal_bin="quarterly")
## sea surface height
# env_bin(sp_trait="pelagic",
#        env_path="F:/cmems/IndoWPac_slh_1m_y2015_y2024.nc",
#        out_dir="F:/whaleshark_sdm/bin",
#        var_name="zos_mean",
#        lon_min=90,
#        lon_max=180,
#        lat_min=-41,
#        lat_max=40,
#        date_min=ymd(20150101),
#        date_max = ymd(20241201),

#        temporal_bin="quarterly")
## mixed layer depth
# env_bin(sp_trait="pelagic",
#        env_path="F:/cmems/IndoWPac_mlt_1m_y2015_y2024.nc",
#        out_dir="F:/whaleshark_sdm/bin",
#        var_name="mlotst_mean",
#        lon_min=90,
#        lon_max=180,
#        lat_min=-41,
#        lat_max=40,
#        date_min=ymd(20150101),
#        date_max = ymd(20241201),

#        temporal_bin="quarterly")
## sea surface salinity
# env_bin(sp_trait="pelagic",
#        env_path="F:/cmems/IndoWPac_so_1m_y2015_y2024.nc",
#        out_dir="F:/whaleshark_sdm/bin",
#        var_name="so_mean",
#        lon_min=90,
#        lon_max=180,
#        lat_min=-41,
#        lat_max=40,
#        date_min=ymd(20150101),
#        date_max = ymd(20241201),

#        temporal_bin="quarterly")
## sea surface NPP
# env_bin(sp_trait="pelagic",
#        env_path="F:/cmems/IndoWPac_npp_1m_y2015_y2024.nc",
#        out_dir="F:/whaleshark_sdm/bin",
#        var_name="nppv",
#        lon_min=90,
#        lon_max=180,
#        lat_min=-41,
#        lat_max=40,
#        date_min=ymd(20150101),
#        date_max = ymd(20241201),

#        temporal_bin="quarterly")
## sea surface dissolved oxygen
# env_bin(sp_trait="pelagic",
#        env_path="F:/cmems/IndoWPac_o2_1m_y2015_y2024.nc",
#        out_dir="F:/whaleshark_sdm/bin",
#        var_name="o2",
#        lon_min=90,
#        lon_max=180,
#        lat_min=-41,
#        lat_max=40,
#        date_min=ymd(20150101),
#        date_max = ymd(20241201),

#        temporal_bin="quarterly")
```
Please note that env_bin() function will create sub-folders (e.g., Q1, Q2, Q3, Q4) for each temporal bin, and save the predictors to the corresponding sub-folders. You may, however, need to rename them to the more meaningful type if necessary. For instance, ocean temperature is usually named as "thetao_mean" in the original cms dataset, and you may change it to "Temperature". 

### 1.2.5 Download the depth dataset 
The Bio-ORACLE provides statistics of ocean depths (mean, maximum, etc.). Here I did not run the code while creating this vignette, but you can run it after removing hash tags. 
```{r download depth, cache=TRUE}
# devtools::install_github("bio-oracle/biooracler")
# library(biooracler)
# library(terra)
# list<-list_layers("Ocean depth")
# dataset_id <- list$dataset_id[158]
# latitude <- c(-41, 40)
# longitude <- c(90, 179.975)
# constraints <- list(latitude, longitude)
# names(constraints) <- c("latitude", "longitude")
# variables <- "bathymetry_mean"
# depth <- download_layers(dataset_id, variables, constraints)
# depth<-abs(depth) # convert values to positive.
# temp<-terra::rast("F:/whaleshark_sdm/bin/Q1/Temperature.asc")
# depth_resampled <- terra::resample(depth, temp, method = "bilinear")
# writeRaster(depth_resampled,"F:/whaleshark_sdm/bin/Depth.asc",
#             NAflag=-9999,overwrite=TRUE,filetype = "AAIGrid")
```
Please note that you need to add the depth_mean.asc sub-folders (e.g., Q1, Q2, Q3, Q4) for each temporal bin, and save the predictors to the corresponding sub-folders. You may, however, need to rename them to the more meaningful type if necessary. For instance, ocean temperature is usually named as "thetao_mean" in the original cms dataset, and you may change it to "temperature". I have uploaded the prepared quarterly predictor datasets to the package, so I will get access to these system files directly as follows. 

### 1.2.6 Plot environmental predictors (optional)
```{r plot_data, cache=TRUE}
# Load required packages
# library(terra)
# library(ggplot2)
# library(cowplot)
# library(dplyr)
# library(scales)   # for squish
# 
# # Main directory and quarters
# main_dir <- system.file("extdata", package = "OceanSDM") # replace to your own directory where the predictor data were stored
# quarters <- c("Q1", "Q2", "Q3", "Q4") # replace to the sub-folder for each quarter under the main_dir
# 
# # Loop over each quarter
# for (q in quarters) {
#   quarter_path <- file.path(main_dir, q)
# 
#   # Read all .asc files in the quarter folder
#   files <- list.files(quarter_path, pattern = "\\.asc$", full.names = TRUE,
#              ignore.case = TRUE)
#   r <- terra::rast(files)
#   names(r) <- tools::file_path_sans_ext(basename(files))
# 
#   n_layers <- nlyr(r)
# 
# ## Convert raster layers to a data frame for ggplot
#   df_list <- list()
#   for (i in 1:n_layers) {
#     df <- as.data.frame(r[[i]], xy = TRUE, na.rm = FALSE)
#     colnames(df) <- c("x", "y", "value")
#     df$variable <- names(r)[i]
#     df_list[[i]] <- df
#   }
#   df_all <- bind_rows(df_list)
# 
# # Create a list of ggplot objects with external labels
#   gg_list <- list()
#   for (i in 1:n_layers) {
#     layer_name <- names(r)[i]
#     df_sub <- filter(df_all, variable == layer_name)
# ## Compute 2% and 98% percentiles for color stretching (percent clip)
#     values_vec <- df_sub$value[!is.na(df_sub$value)]
#     if (length(values_vec) > 0) {
#       p2 <- quantile(values_vec, 0.02, na.rm = TRUE)
#       p98 <- quantile(values_vec, 0.98, na.rm = TRUE)
#       # If p2 == p98, avoid identical limits; use full range
#       if (p2 == p98) {
#         p2 <- min(values_vec, na.rm = TRUE)
#         p98 <- max(values_vec, na.rm = TRUE)
#       }
#     } else {
#       p2 <- 0; p98 <- 1  # fallback
#     }
# 
#     p <- ggplot(df_sub, aes(x = x, y = y, fill = value)) +
#       geom_raster() +
#       scale_fill_viridis_c(
#         option = "inferno",
#         begin = 0.1,
#         end = 0.9,
#         limits = c(p2, p98),      # percent clip range
#         oob = squish,            # squish extremes to the nearest limit color
#         na.value = "transparent"
#       ) +
#       coord_fixed() +
#       labs(title = layer_name, fill = "Value") +
#       theme_minimal() +
#       theme(
#         plot.title = element_text(size =8, face = "bold", hjust = 0.5),
#         legend.position = "right",
#         legend.key.size = unit(0.25, "cm"),
#         legend.text = element_text(size = 4),
#         legend.title = element_text(size = 5),
#         legend.spacing = unit(0.1, "cm"),
#         axis.text = element_blank(),
#         axis.title = element_blank(),
#         axis.ticks = element_blank(),
#         panel.grid = element_blank(),
#         plot.margin = ggplot2::margin(t = 5, r = 5, b = 5, l = 5, unit = "pt")
#       )
# 
#     ggdraw_obj <- ggdraw(p) +
#       draw_label(
#         label = paste0("(", letters[i], ")"),
#         x = 0, y = 1,
#         hjust = -0.05,
#         vjust = 1,
#         size = 8,
#         fontface = "bold"
#       )
#     gg_list[[i]] <- ggdraw_obj
#   }
# 
# ## Arrange subplots
#   ncol <- ceiling(sqrt(n_layers))
#   nrow <- ceiling(n_layers / ncol)
#   combined <- plot_grid(plotlist = gg_list,
#                  ncol = ncol, nrow = nrow, align = "hv")
# 
# ## Create title with "Environmental predictors of QX" and separate by ~2 lines
#   title_text <- paste0("Environmental predictors of ", q)
#   title_gg <- ggdraw() +
#     draw_label(title_text, x = 0.5, y = 0.5, size = 16, fontface = "bold") +
#     theme(plot.margin = ggplot2::margin(b = 20, t = 10, unit = "pt"))
# 
# #   # Combine title and plot vertically
#   final_plot <- plot_grid(title_gg, combined, ncol = 1,
#                   rel_heights = c(0.12, 1), align = "v")
# 
# ## Display the plot
#   print(final_plot)
# }

```
## 1.3 Check occurrence coordinates against species range map with range_check_single() (optional) #############
Sometimes, you may need to check if all your species occurrences fall within the species range map. You can use range_check_single() for a single species, or range_check() for multiple taxa.
```{r range check, cache=FALSE}
# You can get access to the example occurrences of whale sharks in the system file folder extdata.
occ_path <-system.file("extdata", "R_typus.csv", package = "OceanSDM")

# You can get access to the example range map of whale sharks in the system file folder extdata.
range_path <-system.file("extdata", "R_typus_wip2.shp", package = "OceanSDM")
# But for creating this vignette, I loaded them from my PC.
#occ_path<-"F:/whaleshark_sdm/occ/R_typus.csv"
#range_path<-"F:/whaleshark_sdm/range_map/R_typus_wip2.shp"
result <- range_check_single(
  species_name="Rhincodon typus",
      occ_data = occ_path,
      range_path = range_path,
      buffer_km = 100,
  plot = T,
  xlim = c(90, 179.975),
  ylim = c(-41, 40)
  )
occ_range<-rbind(result$valid_points,result$invalid_points) # Note here I kept the 'invalid_points' because they are actually valid seasonal sightings even though they fell beyond the IUCN range map.
# # save the occurrences
write.csv(occ_range,"F:/whaleshark_sdm/occ/R_typus_r.csv",row.names = F)
```
## 1.4 Adjust occurrence coordinates with correct_ob() #############
Previous SDM studies often find that some occurrences were not spatially matched with the environmental layer, but these points fall within a close distance to the environmental layer. This may result from the resolution uncertainty of the occurrences or the coarse resolution of environmental data. I might be worthy correcting them so they match each other, especially when your focal species are naturally rare and you want to keep as many occurrences as possible. Here I showed an example with temperature of the 1st quarter bin
```{r adjust_coordinates, cache=FALSE}

## Define input paths
ev_layer<-system.file("extdata", "Q1/Depth.asc", package = "OceanSDM")
occ_range<-system.file("extdata", "R_typus_r.csv", package = "OceanSDM")
output_path<-"F:/whaleshark_sdm/occ/R_typus_adj.csv"
corrected_ob <- correct_ob(
  occ_path = occ_range,
  predictor_raster_path = ev_layer,
  max_distance_km = 56,
  out_dir = output_path,
  verbose = TRUE, plot=T,
  xlim = c(90, 179.975),
  ylim = c(-41, 40)
)
summary(corrected_ob)

```

## 1.5 Check and remove spatio-temporal bias with bias_check() and bias_reduce()
Spatio-temporal sampling bias is a common issue in global biodiversity data and it degrades model robustness. Here, bias_check() and bias_reduce() are two functions to examine and reduce sampling bias (see details in Supplementary Information), respectively, adapted from the dynamicSDM (Dobson et al., 2023). 

Although designed for highly-mobile marine fish, these data-preparation functions can apply broadly to other marine species.
```{r check & remove bias, cache=TRUE}
library(dplyr)
## Read input data or skip this line if you have run the section 1.4
corrected_ob<-read.csv(system.file("extdata", "R_typus_adj.csv", package = "OceanSDM"))
## Read input data or skip this line if you have run the section 1.4
range_path <-system.file("extdata", "R_typus_wip2.shp", package = "OceanSDM")
colnames(corrected_ob)
occ_ob<-corrected_ob[,c(1:7)] # removed unuseful columns
colnames(occ_ob)[2:3]<-c("y","x") # rename latitude and longitude columns
#### For the quarterly SDMs
occ_ob2 <- occ_ob %>%
  mutate(
         quarter = dplyr::case_when(
           month %in% 1:3   ~ "Q1",
           month %in% 4:6   ~ "Q2",
           month %in% 7:9  ~ "Q3",
           month %in% 10:12 ~ "Q4",
           TRUE ~ NA_character_
         ))%>% dplyr::filter(!is.na(quarter))
table(occ_ob2$quarter)

#### For the seasonal SDMs
# occ_ob3 <- occ_ob2 %>%
#  mutate(date = make_date(year, month, day),
#         adjusted_date = date - days(73), 
# # Parse date information and adjust for sea temperature lag (73 days)
#         adjusted_month = month(adjusted_date),
#         hemisphere = ifelse(y >= 0, "Northern", "Southern"),
#         season = case_when(
#           hemisphere == "Northern" & adjusted_month %in% 3:5   ~ "Spring",
#           hemisphere == "Northern" & adjusted_month %in% 6:8   ~ "Summer",
#           hemisphere == "Northern" & adjusted_month %in% 9:11  ~ "Autumn",
 #          hemisphere == "Northern" & adjusted_month %in% c(12,1,2) ~ "Winter",
  #         hemisphere == "Southern" & adjusted_month %in% 3:5   ~ "Autumn",
  #         hemisphere == "Southern" & adjusted_month %in% 6:8   ~ "Winter",
  #         hemisphere == "Southern" & adjusted_month %in% 9:11  ~ "Spring",
#           hemisphere == "Southern" & adjusted_month %in% c(12,1,2) ~ "Summer",
#           TRUE ~ NA_character_
#         ))
# occ_ob2 <- occ_ob3 %>% filter(!is.na(season))
#range_path<-"F:/whaleshark_sdm/range_map/R_typus_wip2.shp"
## check sampling bias
bias_test<-bias_check(occ_data=occ_ob2,
                        temporal_level="quarter",
                        spatial_method="range_map",
                        range_path=range_path,
                        prj = "+proj=longlat +datum=WGS84")
bias_test$Plots
occ_thinned<-bias_reduce(occ_data=occ_ob2,
                           bias_result=bias_test,
                  min_dist_km = 56,
                          thin_reps = 50,
                          temporal_level = "quarter",
                          seed = 123)
bias_test<-bias_check(occ_data=occ_thinned,  
                        temporal_level="quarter",
                        spatial_method="range_map",
                        range_path=range_path,
                        prj = "+proj=longlat +datum=WGS84")
bias_test$Plots 


#### save the thinned occurrences
write.table(occ_thinned, "F:/whaleshark_sdm/occ/R_typus_qs.txt", 
   append = FALSE, sep = "\t", dec = ".",row.names = FALSE, col.names = TRUE)
occ_q1<-occ_thinned[which(occ_thinned$quarter=="Q1"),]
write.table(occ_q1, "F:/whaleshark_sdm/occ/R_typus_q1.txt", 
 append = FALSE, sep = "\t", dec = ".",row.names = FALSE, col.names = TRUE)
occ_q2<-occ_thinned[which(occ_thinned$quarter=="Q2"),]
write.table(occ_q2, "F:/whaleshark_sdm/occ/R_typus_q2.txt", 
 append = FALSE, sep = "\t", dec = ".",row.names = FALSE, col.names = TRUE)
occ_q3<-occ_thinned[which(occ_thinned$quarter=="Q3"),]
write.table(occ_q3, "F:/whaleshark_sdm/occ/R_typus_q3.txt", 
   append = FALSE, sep = "\t", dec = ".",row.names = FALSE, col.names = TRUE)
occ_q4<-occ_thinned[which(occ_thinned$quarter=="Q4"),]
write.table(occ_q4, "F:/whaleshark_sdm/occ/R_typus_q4.txt", 
   append = FALSE, sep = "\t", dec = ".", row.names = FALSE, col.names = TRUE)
```
Please note that the spatial bias still exist but have been largely reduced. A certain level of spatial autocorrelation of occurrences are ecological meaningful given species are rarely uniformly distributed in the ocean (due to the clumping nature of resources and sometimes species natural schooling or aggregation habits), and such autocorrelation is also allowed for machine-learning algorithms.

# 2. Build 2D SDMs for each temporal bin and generate related figures ####
## 2.1 SDM construction for each quarter with TB_sdm()
To predict horizontal geographic distributions within each temporal bin, we created a single function TB_sdm(), which leveraged some key functions from the packages sdm  (Naimi & Araújo, 2016), blockCV (Valavi et al., 2018), sf (Pebesma, 2018), terra (Hijmans et al., 2022), and Metrics (Hamner & Frasco, 2012). This function streamlines multiple modelling steps: (i) reading occurrences and predictors, (ii) creating folds for spatial block cross-validation, (iii) pseudo absence generation, (iv) collinearity diagnostic and selection of predictors (in an interactive manner), (v) model calibration and evaluation, (vi) ensemble construction and final prediction, and (vii) variable importance and partial dependence data extraction. It constructs models based on the sdm package, which supports classical statistical methods, machine learning algorithms, envelope methods, as well as their ensembles (see details in Naimi & Araújo, 2016). 
Here I only run the code for the 1st quarter as an example to create this vignette webpage.
```{r build 2D-SDM, cache=TRUE}
#### Prior to this step, please rename the environmental predictors to make the
#### the names of variables of each quarterly bin the same as follows:
#### Depth, Mixed_layer_depth, Dissolved_oxygen, Temperature, Salinity, 
####      Net_primary_productivity, Sea_surface_height.
## Q1
rm()
gc()
# You can get access to the Q1 occurrences of whale sharks in the system file folder extdata.
d_occ <- system.file("extdata", "R_typus_q1.txt", package = "OceanSDM")
d_env <- system.file("extdata", "Q1", package = "OceanSDM")
d_prj<-system.file("extdata", "proj", package = "OceanSDM")  # Here I set up a folder where predictors of the 2nd quarter were stored, and was used as a faked future climate scenarios to demonstrate how future projections can be done.
d_res<-"F:/whaleshark_sdm/result_sdm/Q1" # Change to your own directory
#
library(sdm)
sdm_q1<- TB_sdm(env_path=d_env, occ_path=d_occ, res_path=d_res,
                species="Species", x="x", y="y",
                algorithms = c("brt", "rf", "svm"),
                cv_method = "random", n_folds =5,
                pre_abs_ratio = 0.3, block_size = 500000,
                var.importance = TRUE, response.curve = TRUE,
                ensemble.method = "weighted", ensemble.stat = "TSS",
                ncore = 1, seed = 123,
                proj_path = d_prj,
                collinearity_thresh = 0.75, remove_collinear = F)
## Q2
d_occ <- system.file("extdata", "R_typus_q2.txt", package = "OceanSDM")
d_env <- system.file("extdata", "Q2", package = "OceanSDM")
d_res<-"F:/whaleshark_sdm/result_sdm/Q2"
sdm_q2<- TB_sdm(env_path=d_env, occ_path=d_occ, res_path=d_res,
                species="Species", x="x", y="y",
                algorithms = c("brt", "rf", "svm"),
                cv_method = "random", n_folds =5,
                pre_abs_ratio = 0.3, block_size = 500000,
                var.importance = TRUE, response.curve = TRUE,
                ensemble.method = "weighted", ensemble.stat = "TSS",
                ncore = 1, seed = 123,
                collinearity_thresh = 0.75, remove_collinear = F)
# # Q3
d_occ <- system.file("extdata", "R_typus_q3.txt", package = "OceanSDM")
d_env <- system.file("extdata", "Q3", package = "OceanSDM")
d_res<-"F:/whaleshark_sdm/result_sdm/Q3"
sdm_q3<- TB_sdm(env_path=d_env, occ_path=d_occ, res_path=d_res,
                species="Species", x="x", y="y",
                algorithms = c("brt", "rf", "svm"),
                cv_method = "random", n_folds =5,
                pre_abs_ratio = 0.3, block_size = 500000,
                var.importance = TRUE, response.curve = TRUE,
                ensemble.method = "weighted", ensemble.stat = "TSS",
                ncore = 1, seed = 123,
                collinearity_thresh = 0.75, remove_collinear = F)
# # Q4
d_occ <- system.file("extdata", "R_typus_q4.txt", package = "OceanSDM")
d_env <- system.file("extdata", "Q4", package = "OceanSDM")
d_res<-"F:/whaleshark_sdm/result_sdm/Q4"
sdm_q4<- TB_sdm(env_path=d_env, occ_path=d_occ, res_path=d_res,
                species="Species", x="x", y="y",
                algorithms = c("brt", "rf", "svm"),
                cv_method = "random", n_folds =5,
                pre_abs_ratio = 0.3, block_size = 500000,
                var.importance = TRUE, response.curve = TRUE,
                ensemble.method = "weighted", ensemble.stat = "TSS",
                ncore = 1, seed = 123,
                collinearity_thresh = 0.75, remove_collinear = F)
## save data
# save(sdm_q1,sdm_q2,sdm_q3,sdm_q4, file = "F:/whaleshark_sdm/example.RData")

```

## 2.2 Plot predictive maps, variable importance and species response curves with TB_map(), TB_varimp(), and TB_rescur() #
After successful 2D-SDM construction, TB_map(), TB_varimp(), and TB_rescur() automatically produces related figures (i.e., probability and binary maps, variable importance, and species response curves) required for publication.
```{r plot 2D-SDM, cache=FALSE}
## Clean the R environment
rm()
gc()
## Plot predictive maps
map_plot<-TB_map(root_path = "F:/whaleshark_sdm/result_sdm",
  time_type = "quarter",
  output_width = 8,
  output_height = 12,
  output_filename_base = "SDM_quarterly")
print(map_plot)

# Name each temporal bin
time_names <- c("1st Quarter", "2nd Quarter", "3rd Quarter", "4th Quarter")

#  Plot variable importance
imp_plot <- TB_varimp(root_path = "F:/whaleshark_sdm/result_sdm",
                      time_type = "quarter",
                      file_name = "VarImportance_Table.csv",
                      time_names = time_names,
                      method = "AUCtest",
                      colors = c("1st Quarter" = "#FFB347",
                                 "2nd Quarter" = "#4CAF50",
                                 "3rd Quarter" = "#FF6B4A",
                                 "4th Quarter" = "#5D9BEC"),
                      title = NULL,
                      x_label = "Importance",
                      y_label = NULL,
                      group_label= "Quarters",
                      width = 8,
                      height = 6,
                      save_path = "F:/whaleshark_sdm/result_sdm/variable_importance.tif")
print(imp_plot)
#  Plot response curves panel (legend at the bottom)
res_plot <- TB_rescur(root_path = "F:/whaleshark_sdm/result_sdm",
                      time_type = "quarter",
                      file_name = "rcurve_Table.csv",
                      colors = c("1st Quarter" = "#FFB347",
                                 "2nd Quarter" = "#4CAF50",
                                 "3rd Quarter" = "#FF6B4A",
                                 "4th Quarter" = "#5D9BEC"),
                      time_names = time_names,
                      time_folders = NULL,
                      legend_position = "bottom",
                      ncol=3, nrow = NULL,
                      width = 9,
                      height = 5,
                      save_path = "F:/whaleshark_sdm/result_sdm/response_curves.tif")
print(res_plot)

```

# 3. Estimate suitable/preferred depth and map the 3D habitat for each temporal bin
## 3.1 Estimate climatic niche (temperature) with niche_tag() or niche_occ() ##
A key contribution of this package is providing tools to estimate species realized climatic niche based on either tagging studies or species traits about water-column preference. I correspondingly developed two functions, niche_tag() and niche_occ(). Both functions will save a dataframe "summary_niche_fit.csv", which stored the estimated optimal temperature, preferred temperature range, and tolerable temperature range. It will be used as an input for TB_depth(). The niche fitting plot was also saved.

### 3.1.1 Use niche_tag to estimate thermal niche based on temperature utilization data from tagging studies
Many studies have indicated that the observed depths of highly-mobile marine fishes are primarily determined by physiological constraints (especially thermal niche and oxygen demand) (Arrowsmith et al., 2021; Coffey et al., 2017; Dahms & Killen, 2023; Valle et al., 2024), prey availability (Braun et al., 2023), and life history needs (e.g., reproduction, nursery, and wintering) (Andrzejaczek et al., 2019). While prey availability and life history needs are vital, these factors are currently difficult to monitor or correlate with depths. Nevertheless, species realized climatic niche (e.g., the observed thermal niche) is a result of the integrated impacts of such physical and biological factors and can be correlated with depths. The major challenge is thus how to quantify species realized climatic niche.

An optimal approach to estimate species realized climatic niche is using tag-based frequency of occupancy across in-situ ocean climate bins (e.g., temperature bins), which can be readily extracted from figures (using WebPlotDigitizer) in published tagging studies. niche_tag() estimates tolerable climate range (TCR) and preferred climate range (PCR) from such data. It first combines frequency data across studies using a weighted average approach (see the equation in the paper).

The function then allows users to choose suitable approaches to estimate TCR and PCR. For TCR, users can apply specific percentiles (default, 1st and 99th) of the temperature data, so as for PCR (default, 25th and 75th) (Hertz et al., 1993). The function also provides another option—the full width at half maximum (FWHM) from Gaussian kernel density estimation—to estimate PCR.

When tag-based temperature data are unavailable, niche_occ() implements an alternative, trait based approach. It utilizes bias-reduced occurrences to extract real-time temperatures from the depth layer preferred by species (e.g., surface for pelagic fish) rather than depths recorded in biodiversity databases as in the voluModel package. It then allows users to select similar methods (quantiles or FWHM) to estimate TCR and PCR as mentioned for niche_tag().


```{r niche_tag,cache=TRUE}
## Article 1 data (used as the standard bin given the no. of tagged individuals is the highest)
article1_df <- data.frame(
  bin_lower = c(0, 3, 6, 9, 12, 15, 18, 21, 24, 27, 30, 33, 36, 39),
  bin_upper = c(3, 6, 9, 12, 15, 18, 21, 24, 27, 30, 33, 36, 39, 42),
  percentage = c(0,0,0,0,0,0,0.3, 11, 29, 37, 16, 6, 0.4, 0.3) / 100,
  article_weight = 50  # Sample size: 50 individuals
)

## Article 2 data (different binning scheme)
article2_df <- data.frame(
  bin_lower = c(10, 12.5, 15, 17.5, 20, 22.5, 25, 27.5, 30),
  bin_upper = c(12.5, 15, 17.5, 20, 22.5, 25, 27.5, 30, 32.5),
  percentage =c(0.8856, 0.8856, 0.6642, 0.4428, 0.4428, 4.6494, 29.4465, 59.3358, 2.8782)/100 ,
  article_weight = 8  # Sample size: 8 individuals
)

## Article 3 data (another different binning scheme)
article3_df <- data.frame(
  bin_lower = c(10, 12, 14, 16, 18, 20, 22, 24, 26, 28),
  bin_upper = c(12, 14, 16, 18, 20, 22, 24, 26, 28, 30),
  percentage = c(0.65445, 3.02054, 4.36468, 4.36468, 4.36468, 6.207209, 12.58558, 25.50846, 26.01188, 12.91784)/100 ,
  article_weight = 1  # Sample size: 1 individuals
)

## Article 4 data (another different binning scheme)
article4_df <- data.frame(
  bin_lower = c(0, 2, 5, 10, 15, 17.5, 20, 22, 25, 27, 29,31),
  bin_upper = c(2, 5, 10, 15, 17.5, 20, 22, 25, 27, 29,31,33),
  percentage =  c(0.039048, 0.002092, 0.055783, 2.545097, 4.554678, 8.029258, 18.46834, 44.00787, 12.77708, 8.686102, 0.594088, 0.240564)/100,
  article_weight = 15  # Sample size: 15 individuals
)

## Article 5 data (another different binning scheme)
article5_df <- data.frame(
  bin_lower = c(0, 5, 10, 15, 18, 20, 22, 24, 26, 28, 30, 32, 34),
  bin_upper = c(5, 10, 15, 18, 20, 22, 24, 26, 28, 30, 32, 34,36),
  percentage =  c(0.002896, 0.004234, 0.013009, 0.012676, 0.011109, 0.018206, 0.055804, 0.443287, 0.41829, 0.020489,0,0,0),
  article_weight = 27  # Sample size: 15 individuals
)
## Article 6 data (another different binning scheme)
article6_df <- data.frame(
  bin_lower = c(4,8,12,16,20,22,24,26,28),
  bin_upper = c(8,12,16,20,22,24,26,28,30),
  percentage =  c(0.0188, 0.0075, 0.0063, 0.0088, 0.0151, 0.1217, 0.3111, 0.4027, 0.1116),
  article_weight = 17  # Sample size: 15 individuals
)
## Article 7 data (another different binning scheme)
article7_df <- data.frame(
  bin_lower = c(10,12.5,15,17.5,20,22.5,25,27.5,30,32.5),
  bin_upper = c(12.5,15,17.5,20,22.5,25,27.5,30,32.5,35),
  percentage =  c(0, 0.029412, 0.011765, 0.011765, 0.019608, 0.060785, 0.372549, 0.37647, 0.115686, 0.001961),
  article_weight = 17  # Sample size: 15 individuals
)

## Create list of articles
data_list <- list(article1_df, article2_df, article3_df,article4_df,article5_df,article6_df,article7_df)

## Define unified global binning scheme (follow article1_df)
global_bins <- article1_df[,1:2]
## Run estimation
temp_fit<-niche_tag(
    data_list,
    global_bins,
    samples_per_article = 10000,
    toler_quantiles = c(0.01, 0.99),
    pref_method = "FWHM",
    pref_quantiles =NULL,
    variable_name = "Temperature",
    unit = "℃",
    seed = 123,
    width = 6,
    height = 5,
    save_path = "F:/whaleshark_sdm/temp_fit/tag"
)
```


### 3.1.2 Extract temperature data from thinned occurrences to estimate climatic niche
Here, based on the pelagic trait of the focal species, I used niche_occ() to extract the temperature at the depth of 1 m at the thinned occurrences to estimate thermal niche.
```{r niche_occ,cache=FALSE}
occ_path<-system.file("extdata", "R_typus_qs.txt", package = "OceanSDM")

var_name<-"thetao_mean"
# ocean temperature at the surface layer (depth = 1m)
temp_path<-system.file("extdata", "cms/IndoWPac_temp_1m_y2015_y2024.nc", package = "OceanSDM")
save_path <- "F:/whaleshark_sdm/temp_fit/habitat"
min_year<-2015
lon_range<-c(90,180)
lat_range<-c(-41,40)
width<-6
height<-5
temp_occ<-niche_occ(
    occ_path,
    temp_path,
    toler_quantiles = c(0.01, 0.99),
    pref_method = "FWHM",
    pref_quantiles =NULL,
    save_path,
    min_year,
    var_name = "thetao_mean",
    env_label = "Temperature",
    unit = "℃",
    lon_range,
    lat_range,
    width = 8,
    height = 5
)
```

### 3.1.3 Extract oxygen data (optional)
Although the depth limits might be best estimated with thermal niche, other physiological constraints such as demand of oxygen can be estimated as well with data in a user-defined depth at species occurrences. 
```{r oxygen niche,cache=TRUE}
# occ_path<-system.file("extdata", "R_typus_qs.txt", package = "OceanSDM")
# var_name<-"o2"
## Download the O2 data with env_download() first
# o2_path<-"F:/cmems/IndoWPac_o2_1m_y2015_y2024.nc" # dissolved oxygen at the surface layer (depth = 1m)
# save_path <- "F:/whaleshark_sdm/o2_fit/habitat"
# 
# oxy_occ<-niche_occ(
#   occ_path,
#   o2_path,
#   toler_quantiles = c(0.01, 0.99),
#   pref_method = "FWHM",
#   pref_quantiles =NULL,
#   save_path,
#   min_year,
#   var_name = "o2",
#   env_label = "Dissolved oxygen",
#   unit = "μmol/L",
#   lon_range,
#   lat_range,
#   width = 8,
#   height = 5
# )
```

## 3.2 Create temporally-binned temperature for each depth layer ##
Before predicting vertical distributions across space and time, clim_bin() calculates the average climate value at each depth layer of each predicted suitable pixel in each temporal bin (derived from TB_sdm).
```{r clim_bin,cache=FALSE}
## Collect predicted binary maps of each temporal bin (here, quarters)
# quarter_files<-c(
#   Q1="F:/whaleshark_sdm/result_sdm/Q1/predicted_presence.asc",
#   Q2="F:/whaleshark_sdm/result_sdm/Q2/predicted_presence.asc",
#   Q3="F:/whaleshark_sdm/result_sdm/Q3/predicted_presence.asc",
#   Q4="F:/whaleshark_sdm/result_sdm/Q4/predicted_presence.asc")
# presence_asc <- lapply(quarter_files, terra::rast)
# save_path<-"F:/whaleshark_sdm/4D"

## In subsection 1.2.2, you can run the code to downloaded temperature data for three water columns: 0-200m, 200-400m, 400-1050m. And then you can employ clim_bin() to create temporally-binned predictors for each depth layer based on these three .nc files. 
# temp_nc<-"F:/whaleshark_sdm/cmems/IndoWPac_temp_200m_y2015_y2024.nc"
# clim_bin(temp_nc, presence_asc, time_type = "quarter",
#          save_path=save_path, hemisphere = "north", na_value = -9999)
# temp_nc<-"F:/whaleshark_sdm/cmems/IndoWPac_temp_200_400m_y2015_y2024.nc"
# clim_bin(temp_nc, presence_asc, time_type = "quarter",
#          save_path=save_path,  na_value = -9999)
# temp_nc<-"F:/whaleshark_sdm/cmems/IndoWPac_temp_400_1050m_y2015_y2024.nc"
# clim_bin(temp_nc, presence_asc, time_type = "quarter",
#          save_path=save_path, na_value = -9999)
```

## 3.3 Calculate TD and PD for each temporal bin with TB_depth() ######
TB_depth() calculates tolerable depths (TD) and preferred depths (PD) in each pixel for each temporal bin by comparing TCR and PCR, respectively, with the climate value along the vertical gradient. It outputs three statistics (minimum, maximum, range) for TD and PD and plots the vertical climate profile for a random sample (default, 100) of grids. However, this process is very time-consuming (lasting > 8 hrs) given the function has to calculate these statistics grid by grid and there are 47 depth layers. Therefore, I did not run the following code when generating this vignette. But I have stored the outputs while producing the paper.
```{r TB_depth(),cache=TRUE}
##### Example shown with estimates based on tag data
df<-read.csv(system.file("extdata","summary_niche_fit.csv",package = "OceanSDM"))
PTR<-df[2:3,2] # preferred temperature range
TTR<-df[4:5,2] # tolerable temperature range
## Run the whole code in subsection 3.2 to create temporally-binned mean temperature for each depth layer. Here I stored all these files in "F:/whaleshark_sdm/4D"
temp_bin_path<-"F:/whaleshark_sdm/4D"
save_path_TPD<-"F:/whaleshark_sdm/4D/TPD" ## for storing data of Preferred Depth (PD)
save_path_TTD<-"F:/whaleshark_sdm/4D/TTD" ## for storing data of Tolerable Depth (TD)
Sys.setenv(TMPDIR = "F:/temp")   # define a temporal file directory
dir.create("F:/temp", recursive = TRUE, showWarnings = FALSE)
## For Tolerable Depth
TTD_quarter<-TB_depth(temp_bin_path, TTR, time_type = "quarter",
                      layer_option="top",
                      save_path_TTD, na_value = -9999,
                      use_makima = TRUE,
                      N_cells = 50)
## For Preferred Depth (PD), run the following code:
# TPD_quarter<-TB_depth(temp_bin_path, PTR, time_type = "quarter",
#                       layer_option="top",
#                       save_path_TPD, na_value = -9999,
#                       use_makima = TRUE,
#                       N_cells = 50)
```

## 3.4 Plot habitat volume and area with plot_stat_4D() ##############
plot_stat_4D() maps the statistics (Max, Min, Range) of TD and PD across predicted 2D presence maps, generates boxplots, and tests for significant differences among temporal bins using Kruskal Wallis followed by Dunn’s post-hoc tests. It also create a figure to showcase the total suitable habitat volume based on the TD or PD, and estimates total suitable habitat area based on the 2D presence map of each temporal bin. 
```{r plot_stat_4D,cache=FALSE}
## Run the whole code in subsection 3.3 to estimate three parameters (minimum, maximum, range) of thermally preferred depth or thermally tolerable depth. Here I load the input data I already prepared.

TPD_quarter<-list(
"F:/whaleshark_sdm/4D/TPD/Q1/Q1_min_depth.asc",
  "F:/whaleshark_sdm/4D/TPD/Q1/Q1_max_depth.asc",
  "F:/whaleshark_sdm/4D/TPD/Q1/Q1_depth_range.asc",
  "F:/whaleshark_sdm/4D/TPD/Q2/Q2_min_depth.asc",
  "F:/whaleshark_sdm/4D/TPD/Q2/Q2_max_depth.asc",
  "F:/whaleshark_sdm/4D/TPD/Q2/Q2_depth_range.asc",
  "F:/whaleshark_sdm/4D/TPD/Q3/Q3_min_depth.asc",
  "F:/whaleshark_sdm/4D/TPD/Q3/Q3_max_depth.asc",
  "F:/whaleshark_sdm/4D/TPD/Q3/Q3_depth_range.asc",
  "F:/whaleshark_sdm/4D/TPD/Q4/Q4_min_depth.asc",
  "F:/whaleshark_sdm/4D/TPD/Q4/Q4_max_depth.asc",
  "F:/whaleshark_sdm/4D/TPD/Q4/Q4_depth_range.asc"
)
TTD_quarter<-list(
  "F:/whaleshark_sdm/4D/TTD/Q1/Q1_min_depth.asc",
  "F:/whaleshark_sdm/4D/TTD/Q1/Q1_max_depth.asc",
  "F:/whaleshark_sdm/4D/TTD/Q1/Q1_depth_range.asc",
  "F:/whaleshark_sdm/4D/TTD/Q2/Q2_min_depth.asc",
  "F:/whaleshark_sdm/4D/TTD/Q2/Q2_max_depth.asc",
  "F:/whaleshark_sdm/4D/TTD/Q2/Q2_depth_range.asc",
  "F:/whaleshark_sdm/4D/TTD/Q3/Q3_min_depth.asc",
  "F:/whaleshark_sdm/4D/TTD/Q3/Q3_max_depth.asc",
  "F:/whaleshark_sdm/4D/TTD/Q3/Q3_depth_range.asc",
  "F:/whaleshark_sdm/4D/TTD/Q4/Q4_min_depth.asc",
  "F:/whaleshark_sdm/4D/TTD/Q4/Q4_max_depth.asc",
  "F:/whaleshark_sdm/4D/TTD/Q4/Q4_depth_range.asc"
)
save_path_TPD<-"F:/whaleshark_sdm/4D/TPD" ## for Preferred Depth (PD)
save_path_TTD<-"F:/whaleshark_sdm/4D/TTD" ## for Tolerable Depth (TD)
quarter_files<-c(
  Q1="F:/whaleshark_sdm/result_sdm/Q1/predicted_presence.asc",
  Q2="F:/whaleshark_sdm/result_sdm/Q2/predicted_presence.asc",
  Q3="F:/whaleshark_sdm/result_sdm/Q3/predicted_presence.asc",
  Q4="F:/whaleshark_sdm/result_sdm/Q4/predicted_presence.asc")

result_TPD <- plot_stat_4D(
  file_list = TPD_quarter,
  save_path = save_path_TPD,
  time_type = "quarter",
  presence_files = quarter_files
)
result_TPD$depth_range
result_TPD$depth_min
result_TPD$depth_max
result_TPD$overlay_plot
result_TTD <- plot_stat_4D(
  file_list = TTD_quarter,
  save_path = save_path_TTD,
  time_type = "quarter",
  presence_files = quarter_files
)
result_TTD$depth_range
result_TTD$depth_min
result_TTD$depth_max
result_TTD$overlay_plot
```

#################### The end ############################

# References

Andrade, A. F. A. de, Velazco, S. J. E., & De Marco Júnior, P. (2020). ENMTML: An R package for a straightforward construction of complex ecological niche models. Environmental Modelling & Software, 125, 104615. https://doi.org/10.1016/j.envsoft.2019.104615

Andrzejaczek, S., Gleiss, A. C., Pattiaratchi, C. B., & Meekan, M. G. (2019). Patterns and drivers of vertical movements of the large fishes of the epipelagic. Reviews in Fish Biology and Fisheries, 29(2), 335–354. https://doi.org/10.1007/s11160-019-09555-1

Arrowsmith, L., Sequeira, A., Pattiaratchi, C., & Meekan, M. (2021). Water temperature is a key driver of horizontal and vertical movements of an ocean giant, the whale shark Rhincodon typus. Marine Ecology Progress Series, 679, 101–114. https://doi.org/10.3354/meps13899

Braun, C. D., Della Penna, A., Arostegui, M. C., Afonso, P., Berumen, M. L., Block, B. A., Brown, C. A., Fontes, J., Furtado, M., Gallagher, A. J., Gaube, P., Golet, W. J., Kneebone, J., Macena, B. C. L., Mucientes, G., Orbesen, E. S., Queiroz, N., Shea, B. D., Schratwieser, J., … Thorrold, S. R. (2023). Linking vertical movements of large pelagic predators with distribution patterns of biomass in the open ocean. Proceedings of the National Academy of Sciences, 120(47), e2306357120. https://doi.org/10.1073/pnas.2306357120

Coffey, D. M., Carlisle, A. B., Hazen, E. L., & Block, B. A. (2017). Oceanographic drivers of the vertical distribution of a highly migratory, endothermic shark. Scientific Reports, 7(1), 10434. https://doi.org/10.1038/s41598-017-11059-6

Dahms, C., & Killen, S. S. (2023). Temperature change effects on marine fish range shifts: A meta-analysis of ecological and methodological predictors. Global Change Biology, 29(16), 4459–4479. https://doi.org/10.1111/gcb.16770

Dobson, R., Challinor, A. J., Cheke, R. A., Jennings, S., Willis, S. G., & Dallimer, M. (2023). dynamicSDM: An R package for species geographical distribution and abundance modelling at high spatiotemporal resolution. Methods in Ecology and Evolution, 14(5), 1190–1199. https://doi.org/10.1111/2041-210X.14101

Hertz, P. E., Huey, R. B., & Stevenson. (1993). Evaluating Temperature Regulation by Field-Active Ectotherms: The Fallacy of the Inappropriate Question | The American Naturalist: Vol 142, No 5. The American Naturalist, 142(5). https://www.journals.uchicago.edu/doi/10.1086/285573

Naimi, B., & Araújo, M. B. (2016). sdm: A reproducible and extensible R platform for species distribution modelling. Ecography, 39(4), 368–375. https://doi.org/10.1111/ecog.01881

Owens, H. L., & Rahbek, C. (2023). voluModel: Modelling species distributions in three-dimensional space. Methods in Ecology and Evolution, 14(3), 841–847. https://doi.org/10.1111/2041-210X.14064

Valle, M., Ramírez-Romero, E., Ibaibarriaga, L., Citores, L., Fernandes-Salvador, J. A., & Chust, G. (2024). Pan-Atlantic 3D distribution model incorporating water column for commercial fish. Ecological Modelling, 490, 110632. https://doi.org/10.1016/j.ecolmodel.2024.110632
